---
title: "Where `fit_lppls()` spends its time"
subtitle: "The linear-parameter solver, the symengine dependency, Rcpp, and parallelism"
author: "lpplsF"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: kable
params:
  recompute: false   # FALSE (default): present the recorded measurements below.
                     # TRUE: re-run every benchmark (~30 min, machine-specific).
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
RECOMPUTE <- isTRUE(params$recompute)
suppressMessages(pkgload::load_all("..", quiet = TRUE))
```

# Why this document exists

`fit_lppls()` is the workhorse of the package and it is not fast. A single
calibration of the S&P 500 example takes well over a minute, a 101-window
rolling calibration takes a quarter of an hour, and the full thesis-figure
regeneration takes twenty minutes. The question this document answers is where
that time actually goes, and what could be done about it.

The investigation began from a specific suspicion: `create_beta_calculator()`
depends on **symengine**, a computer-algebra system, and symbolic evaluation is
usually slower than a numeric routine doing the same job. That suspicion turns
out to be *largely wrong* — symengine is a minor contributor — but pursuing it
uncovered three larger findings, one of which has nothing to do with the solver
at all.

Everything below is measured, not estimated. Where a number is a projection
rather than a measurement it is labelled as such. One earlier attempt at the
combined measurement produced a self-contradictory result; rather than discard
it silently, section "A measurement that was wrong" explains what was wrong
with it, because the reason turned out to be the most valuable finding in the
document.

## What came of it

This document began as an investigation and has since driven changes to the
package, so the variants below are no longer hypothetical. Three outcomes:

* **The four solvers are shipped**, selectable through
  `fit_lppls(..., beta_method =)`. `"crossprod"` is the default;
  `"symengine"` --- the thesis implementation --- is retained unchanged.
* **The computer-algebra approach became a test oracle rather than a runtime
  path**, for reasons measured in section "The CAS as a test oracle".
* **`factr` and Rcpp remain open questions**, with the evidence set out in
  the sections "A measurement that was wrong" and "Would Rcpp help?".

Section "Recommendations and status" tracks what is done and what is not.

## The machine

All timings are from one machine and are not portable claims:

| | |
|---|---|
| R | 4.5.2 (2025-10-31) |
| platform | macOS, Apple Silicon (arm64) |
| cores | 10 (performance + efficiency, which matters in section "Parallelism") |
| compiler | Apple clang 17.0.0 |
| BLAS/LAPACK | R's reference `libRlapack` (not a tuned BLAS) |

The last row is worth noting: several variants below lean on `crossprod()`, and a
tuned BLAS such as Accelerate or OpenBLAS would change their numbers — probably
in their favour. The reference BLAS is therefore the conservative baseline.

# What is actually being computed

Given the nonlinear parameters $(t_c, m, \omega)$, the LPPLS model is **linear**
in $A, B, C_1, C_2$. This is the Filimonov--Sornette reformulation, and it is the
reason the calibration is tractable at all: the four linear parameters never need
to be searched over, because for any candidate $(t_c, m, \omega)$ they follow in
closed form from ordinary least squares.

Writing $\tau = t_c - t$, the design matrix is

$$
X = \bigl[\, \mathbf{1},\;\; \tau^m,\;\; \tau^m \cos(\omega \log \tau),\;\;
            \tau^m \sin(\omega \log \tau) \,\bigr]
$$

and the solution is $\beta = (X'X)^{-1} X'y$ with $y$ the log-price vector.

Every variant in this document computes exactly that $\beta$. They differ only in
**how**, and the differences are what the benchmark measures. Since the outer
optimiser evaluates this quantity once per objective evaluation — tens of times
per candidate $t_c$, for hundreds of candidates — its cost is multiplied by a
very large number before it reaches the user.

## How the symengine method does it

The `"symengine"` method --- the original implementation, and the one the
thesis used --- has `create_beta_calculator()` use symengine to solve a **general symbolic**
$4 \times 4$ system once, at construction, and compiles four `DoubleVisitor`
closures — one per coefficient — that evaluate the resulting closed-form
expressions. At call time it assembles the 20 distinct entries of $X'X$ and $X'y$
using the `sum_*` helpers in `R/helpers.R`, then evaluates the four expressions.

Two properties of that arrangement matter for what follows.

First, **each `sum_*` helper recomputes the basis from scratch.** `sum_fg()`
calls `lppls_f()` and `lppls_g()`, `sum_fh()` calls `lppls_f()` and `lppls_h()`,
and so on; none of them shares work with any other. Since `lppls_g()` and
`lppls_h()` each independently recompute $\tau^m$ and $\log \tau$ as well,
$\tau^m$ ends up being evaluated roughly 24 times per call, and $\cos$ and $\sin$
roughly 8 times each, where once would do.

Second, **the symmetric entries are computed twice.** The calculator populates
`xx12` and `xx21` with two separate `sum_f()` calls, `xx13` and `xx31` with two
`sum_g()` calls, and so on, even though $X'X$ is symmetric by construction.

# The variants

Six implementations were compared. Four of them now ship as `beta_method`
choices; the remaining two exist purely as controls, to separate the two
candidate explanations for the symengine method's cost. The controls are the reason this document can attribute
the cost rather than merely report it.

**A --- `"symengine"`.** The thesis implementation, retained unmodified.

**B --- shared basis + symengine** (control, not shipped). The *same compiled
symbolic closures*, retrieved from the symengine calculator's own environment via
`environment(bc)$A_func`, but fed from a basis computed once. Nothing about the
symbolic machinery changes; only the redundancy is removed.

**C --- shared basis + `solve()`** (control, not shipped). The same 20 entries, assembled into a
$4 \times 4$ matrix and passed to `solve()`. Nothing about the redundancy
changes; only the symbolic evaluation is replaced.

B and C differ from one another in exactly one respect, which is what makes the
attribution possible: **A versus B isolates the redundant recomputation, and B
versus C isolates the symbolic evaluation.**

**D --- `"crossprod"` (the default).** Abandons the hand-written sums entirely:
build the $n \times 4$ design matrix and let `crossprod()` form $X'X$ and $X'y$
in one BLAS call each.

**E --- `"qr"`.** R's internal least-squares routine `.lm.fit()`, which applies a
Householder QR factorisation to $X$ directly and never forms $X'X$. This is the
numerically conservative choice, for reasons taken up in section
"Is the normal-equations shortcut safe?".

**F --- `"chol"`.** Exploits the fact that $X'X$ is symmetric positive definite,
using `chol()` and `chol2inv()` rather than a general LU solve.

```{r variants, eval=RECOMPUTE}
basis <- function(t, tc, m, omega) {
  tau <- tc - t; taum <- tau^m; lt <- omega * log(tau)
  list(f = taum, g = taum * cos(lt), h = taum * sin(lt))
}
# D, now the shipped default, in full:
beta_crossprod <- function(log_p, t, tc, m, omega) {
  b <- basis(t, tc, m, omega)
  X <- cbind(1, b$f, b$g, b$h)
  s <- solve(crossprod(X), crossprod(X, log_p))
  c(A = s[1], B = s[2], C1 = s[3], C2 = s[4])
}
```

## Results

Median microseconds per call, 200 replications per cell:

| variant | n=200 | n=1000 | n=2428 | speedup at n=2428 |
|---|---:|---:|---:|---:|
| A `"symengine"` | 117.5 | 326.2 | 703.7 | 1.0× |
| B shared + symengine | 43.4 | 65.4 | 105.7 | **6.7×** |
| C shared + `solve` | 18.6 | 41.5 | 82.5 | 8.5× |
| **D `"crossprod"` (default)** | 12.8 | 32.2 | **66.2** | **10.6×** |
| E `"qr"` | 11.0 | 40.5 | 85.1 | 8.3× |
| F `"chol"` | 11.8 | 31.3 | 65.9 | 10.7× |

## The attribution, which is the point

Read the first three rows together:

- **A → B is 6.7×, with symengine still fully in place.** Removing nothing but
  the redundant recomputation of the basis recovers roughly two-thirds of the
  total available gain.
- **B → C is a further 1.3×.** That is the entire contribution of replacing
  symbolic evaluation with a numeric solve.

So the framing that motivated this investigation — that symengine is the
bottleneck — is **not supported**. symengine accounts for about a quarter of the
achievable improvement. The dominant cost is that `sum_f()`, `sum_g()`,
`sum_h()` and their products recompute the same three length-$n$ vectors over
and over within a single call.

This distinction is not academic. It means that a change preserving symengine but
sharing the basis would capture most of the win, if there were a reason to keep
the dependency. Section "The dependency question" argues there is not.

Note also that A and B agree to *exactly* zero relative difference, because B
evaluates the identical symbolic expressions. That is a consistency check on the
harness, not evidence of superior precision.

# Precision

The table above reports each variant's deviation from the shipped calculator.
That is an *agreement* measurement, not an accuracy one: it silently treats the
symengine implementation as ground truth. Since one of the questions this
document exists to answer is whether symengine buys any accuracy, that
assumption has to be removed.

So precision is measured against a **constructed ground truth**. Pick
$\beta_{\text{true}}$, form $X$, and set $y = X\beta_{\text{true}}$ exactly. Every
variant should then return $\beta_{\text{true}}$, and its deviation is its
numerical error rather than its disagreement with a peer.

Conditioning is swept through $m$. As $m \to 0$, $\tau^m \to \mathbf{1}$ and the
second column of $X$ collides with the intercept, so $\kappa$ grows and the
variants can be separated. Maximum relative error in $\beta$:

| m | $\kappa(X)$ | $\kappa(X'X)$ | A symengine | C sums+solve | D crossprod | E QR | F chol |
|---:|---:|---:|---:|---:|---:|---:|---:|
| 0.900 | 2.3e3 | 5.3e6 | 3.7e-14 | 3.3e-14 | 3.0e-14 | **5.8e-16** | 2.5e-14 |
| 0.600 | 3.2e2 | 1.0e5 | 2.7e-13 | 2.7e-13 | 5.0e-14 | **1.3e-15** | 4.3e-14 |
| 0.300 | 6.1e1 | 3.7e3 | 1.5e-12 | 1.4e-12 | 1.7e-12 | **4.6e-15** | 1.9e-12 |
| 0.100 | 4.8e1 | 2.3e3 | 1.4e-10 | 1.4e-10 | 2.1e-12 | **9.4e-15** | 5.5e-12 |
| 0.050 | 7.9e1 | 6.3e3 | 2.8e-11 | 8.9e-12 | 6.3e-11 | **4.4e-14** | 1.5e-11 |
| 0.020 | 1.9e2 | 3.5e4 | 8.7e-09 | 8.8e-09 | 1.1e-09 | **4.1e-14** | 1.1e-10 |
| 0.010 | 3.7e2 | 1.4e5 | 2.7e-08 | 2.7e-08 | 7.3e-12 | **5.1e-14** | 1.4e-09 |
| 0.005 | 7.3e2 | 5.4e5 | 1.9e-07 | 1.9e-07 | 8.4e-09 | **1.0e-13** | 1.5e-08 |

Repeating with realistic noise added to $y$ — where no exact $\beta$ exists, so
the QR solution serves as the reference — reproduces the same ordering.

## What this shows

**QR is far and away the most accurate, and it barely degrades.** Across a
conditioning range spanning three orders of magnitude its error moves only from
$5.8 \times 10^{-16}$ to $1.0 \times 10^{-13}$, because it never forms $X'X$ and
so never squares the condition number.

**symengine is the *least* accurate**, together with the numerically equivalent
`solve()` on the same hand-computed sums. At $m = 0.005$ it is off by
$1.9 \times 10^{-7}$ — six orders of magnitude worse than QR.

**The loss is in forming $X'X$ from the `sum_*` helpers, not in the symbolic
step.** A and C track each other to within a few percent at every conditioning
level, and they differ *only* in whether the $4 \times 4$ system is solved
symbolically or numerically. So the symbolic solve is neither better nor worse
than `solve()`; what costs accuracy is accumulating the 20 entries as R-level
`sum(f * g)` products. `crossprod()` (variant D) forms the same $X'X$ but
accumulates it in BLAS, and is one to four orders of magnitude more accurate for
it.

A residual check confirms the errors are real rather than an artefact of the
comparison. At $m = 0.01$:

| variant | rel. error in $\beta$ | $\lVert y - X\hat\beta \rVert$ |
|---|---:|---:|
| A symengine | 2.65e-08 | 1.93e-10 |
| C sums+solve | 2.71e-08 | 1.97e-10 |
| D crossprod | 7.29e-12 | 1.09e-13 |
| E QR | 5.05e-14 | 1.12e-13 |
| F chol | 1.40e-09 | 4.72e-09 |

symengine's larger parameter error corresponds to a genuinely worse fit — its
residual norm is three orders of magnitude above QR's.

## A correction

An earlier version of this document concluded that "the fastest variant is also
the most faithful, so the usual speed-versus-stability trade-off does not
arise." That was measured as agreement with the shipped calculator, which is
circular: the shipped calculator is itself the least accurate variant, so
matching it closely is not evidence of accuracy. Against a real ground truth the
ordinary trade-off does exist --- QR is meaningfully more accurate than
`crossprod` --- it is simply irrelevant at the parameter values the package
actually admits, for the reason given next.

# Is the normal-equations shortcut safe?

Variants D and F solve the **normal equations**, forming $X'X$ explicitly. This
squares the condition number: $\kappa(X'X) = \kappa(X)^2$. Where $X$ is badly
conditioned the shortcut can lose a damaging number of digits, which is why
`.lm.fit()` (variant E) exists as the conservative option.

Whether that matters is an empirical question about *this* design matrix, so it
was measured rather than assumed:

| case | $\kappa(X)$ | $\kappa(X'X)$ |
|---|---:|---:|
| typical ($n=1000$, $m=0.6$, $\omega=8$) | 1.0e2 | 1.7e4 |
| SPX-like ($n=2428$, $m=0.9$, $\omega=6.16$) | 1.6e3 | 5.8e6 |
| near-degenerate ($m=0.9$, $\omega=13$) | 2.1e3 | 7.0e6 |

Double precision carries roughly 16 significant digits. The worst case here,
$\kappa(X'X) \approx 7 \times 10^6$, costs about 7 of them and leaves 9 — far
more than this application needs, given that the parameters are interpreted to
three or four decimals.

**The package's own bounds are what make this safe.** `fit_lppls()` defaults to
`lower[1] = 0.1`, and Filimonov's admissible range is $0.1 \le m \le 0.9$. Every
row of the precision table at $m \ge 0.1$ has D erring by at most
$2.1 \times 10^{-12}$ — utterly negligible against parameters interpreted to
three or four decimals. The degradation only appears at $m < 0.1$, which the
default search box excludes.

So the choice between D and E is not about correctness within the supported
range; both are exact there for any practical purpose. It is about how much
margin to keep:

- **D (`crossprod`)** is 1.3× faster and adequate for every $m$ the package
  admits by default.
- **E (`.lm.fit`)** costs 30% more, is 100--1000× more accurate, and holds up if
  a user widens `lower` toward $m \to 0$ or fits a series where the power-law and
  constant columns are nearly collinear.

Both ship. **`"crossprod"` is the default**, chosen for speed on the reasoning
that the package's own bounds already exclude the region where it degrades, and
**`"qr"` is one argument away** for anyone who widens them. Neither is a
correctness risk inside the supported range; the decision is only about how much
margin to keep outside it.

# Where the time goes in a whole calibration

A fast solver is only worth having if the solver is where the time is. Profiling
a complete `fit_lppls()` call settles that:

```
=== profile of one fit_lppls() call (shipped) ===
total sampled time: 11.05 s
  "beta_calculator"    9.28 s  (83.9% of total)
```

**83.9% of the entire calibration is inside the beta calculator.** Everything
else — the L-BFGS-B machinery, `eval_lppls()`, table assembly, the final Brent
refinement of $t_c$ — shares the remaining sixth.

Substituting variant D at the namespace level, with no other change, on two real
workloads:

| workload | `"symengine"` | `"crossprod"` (default) | `"qr"` |
|---|---:|---:|---:|
| SPX-scale F2 ($n=2428$, $fh=500$) | 84.40 s | **14.39 s (5.9×)** | 16.99 s (5.0×) |
| MPL mode ($n=1000$, $fh=100$) | 13.17 s | **2.64 s (5.0×)** | 3.06 s (4.3×) |

The estimates agree: $t_c$ to 11 significant digits, SSE to 10 decimals.

The end-to-end factor (5.9×) is smaller than the component factor (10.6×) exactly
as Amdahl's law requires, since a sixth of the work is untouched. This is the
honest figure to quote for the change in isolation.

# Would Rcpp help?

Variant D still pays costs that are unavoidable in vectorised R. It allocates
three length-$n$ vectors for the basis, materialises an $n \times 4$ design
matrix, and passes over the data several times. A compiled implementation can
**fuse** the whole thing: a single loop that computes $\tau^m$, $\cos$ and $\sin$
once per observation and accumulates the 13 distinct inner products directly,
never allocating anything of length $n$ at all.

The prototype is plain Rcpp — no RcppArmadillo — with a hand-written
$4 \times 4$ Gauss--Jordan solve, so the measurement reflects the lightest
possible dependency (`LinkingTo: Rcpp`).

| implementation | µs per call (n=2428) | vs shipped | vs best R |
|---|---:|---:|---:|
| shipped (symengine) | 728.2 | 1.0× | 0.1× |
| R `crossprod` + `solve` | 69.9 | 10.4× | 1.0× |
| **Rcpp, fused single pass** | **32.9** | **22.1×** | **2.1×** |

So Rcpp roughly doubles the best pure-R result, for a total of 22× on the
component.

## The larger Rcpp opportunity is elsewhere

`compute_X_matrix()` and `compute_H_matrix()`, used by the modified profile
likelihood, are **explicit R `for` loops over $n$**, executed once per candidate
$t_c$. They are far more exposed to compilation than the solver ever was:

| routine | R | Rcpp | speedup |
|---|---:|---:|---:|
| `compute_X_matrix()` | 514.6 µs | 37.0 µs | **13.9×** |
| `compute_H_matrix()` | 1110.5 µs | 33.2 µs | **33.4×** |

These are the largest single-routine gains in the document. They matter
specifically for `mode = "MPL"`, which is what the rolling calibrations and every
likelihood interval in the thesis depend on.

## What adopting Rcpp would cost

The package is currently **pure R with no `src/` directory**. That is a real
property worth pricing before giving it up:

- installation from source would require a C++ toolchain on the user's machine;
- `R CMD check` gains a compilation step, and CRAN checks across more platforms;
- `pkgload::load_all()` acquires a compile step, slowing the edit--test loop;
- the package becomes harder to inspect for a reader who knows R but not C++,
  which matters for something written as a companion to a thesis.

The gains are real but second-order relative to the pure-R change: 22× versus
10.6× on the solver. Since symengine is being kept (section
"The dependency question"), the toolchain requirement exists either way,
so that particular objection no longer applies --- but the sequencing argument
does: the cheap change was worth taking first.

# Parallelism

`fit_lppls()` in `"F2"` and `"MPL"` mode loops over `fh` candidate values of
$t_c$, running an independent L-BFGS-B optimisation over $(m, \omega)$ for each.
Those iterations do not communicate. This is the natural parallel axis, and it is
wide: 500 candidates for the SPX calibration.

Measured on the real inner problem, over the already-optimised solver, so that
the figures reflect achievable rather than flattering speedups:

| scheme | time | speedup | efficiency | results identical |
|---|---:|---:|---:|---|
| serial | 22.35 s | 1.00× | — | — |
| `mclapply`, 2 cores | 10.73 s | 2.08× | 104% | yes |
| `mclapply`, 4 cores | 6.33 s | 3.53× | 88% | yes |
| `mclapply`, 8 cores | 4.70 s | 4.76× | 59% | yes |
| `mclapply`, 10 cores | 4.53 s | 4.94× | 49% | yes |
| `parLapply` (PSOCK), 8 | 5.07 s | 4.41× | 55% | yes |

Efficiency falls away sharply beyond four cores. This is an Apple Silicon machine
with a mix of performance and efficiency cores, and the efficiency cores
contribute far less per thread; on a homogeneous many-core machine the curve
would look better. The practical reading is that **the first four cores are worth
having and the rest are not**, at least here.

## The caveats are substantial

Parallelising this loop is not a drop-in change, and three obstacles deserve
weighing before anyone attempts it.

**The loop is not currently side-effect free.** It accumulates
`out_of_range_tracker` and builds `fit2_best_for_each_tc` incrementally. A
parallel version must return per-$t_c$ results and combine them afterwards, which
is a genuine restructuring of `fit_lppls()`, not an annotation.

**Random restarts break reproducibility.** With `num_searches > 1`, each
candidate draws random initial values. Under `mclapply` the RNG stream must be
managed explicitly (`mc.set.seed`, or L'Ecuyer streams) or results become
non-reproducible — and even managed correctly they will not match the serial
sequence. Every cached fit and every thesis figure would shift. The measurements
above used a single deterministic start, which is why results came out identical;
that guarantee does **not** carry over to the multi-start case.

**Nested parallelism is a live hazard in this project.** The rolling and
Monte-Carlo experiments already run `parallel::mclapply(mc.cores = 8)` at the
*experiment* level. If `fit_lppls()` also parallelised internally, the two would
compose into 64 processes on a 10-core machine. Oversubscription of exactly this
kind previously caused a multi-hour stall with no completed work. Any internal
parallelism therefore needs to be **off by default** and controlled by an
explicit argument, so that the outer loop remains the place where cores are
spent.

Given that the outer level is already parallelised and is the coarser, safer
axis, internal parallelism is the least attractive of the three changes
considered here despite its headline number.

# A measurement that was wrong

The first attempt to measure the combined effect produced this:

```
serial, symengine (as shipped) :    2.36 s    1.00x
serial, R crossprod            :   22.43 s    0.11x
```

The shipped implementation appeared **ten times faster** than a variant that had
just beaten it tenfold, and an equality check across variants failed. Rather than
report it, the cause was traced — and it is the most consequential finding here.

The harness ran a real `optim()` call per candidate. The three solvers agree only
to about $10^{-14}$, and those last-digit differences are enough to send L-BFGS-B
down different paths. The variants were therefore not doing equal work: the
timings compared **different numbers of objective evaluations**, not different
speeds.

The reason the paths diverged so wildly is `factr`.

## `factr = 1e-8` asks for the impossible

`optim()`'s `factr` for L-BFGS-B is a multiplier on machine epsilon, not an
absolute tolerance. The package's calls — and every call in the thesis code —
use `factr = 1e-08`, which requests a convergence tolerance of

$$
10^{-8} \times \varepsilon_{\text{mach}}
  = 10^{-8} \times 2.22 \times 10^{-16}
  = 2.22 \times 10^{-24},
$$

eight orders of magnitude below what double precision can represent. That
criterion can never be satisfied. The optimiser therefore does not converge; it
runs until the line search fails and returns **convergence code 52**
(`ABNORMAL_TERMINATION_IN_LNSRCH`).

At a single representative $t_c$, with the identical optimum found either way:

| solver | factr | evaluations | convergence | objective |
|---|---:|---:|---:|---:|
| symengine | 1e-08 | 58 | 52 | 0.26053824 |
| R crossprod | 1e-08 | 62 | 0 | 0.26053824 |
| Rcpp | 1e-08 | 105 | 52 | 0.26053824 |
| symengine | 1e+07 | 8 | 0 | 0.26053824 |
| R crossprod | 1e+07 | 8 | 0 | 0.26053824 |
| Rcpp | 1e+07 | 8 | 0 | 0.26053824 |

With R's default `factr = 1e7` all three agree on 8 evaluations, converge
cleanly, and land on the same objective value.

## Confirming it across the grid

One point proves nothing, so the comparison was repeated over all 500 candidate
$t_c$:

| | `factr = 1e-8` | `factr = 1e+7` |
|---|---:|---:|
| objective evaluations, median | 59 | **9** |
| objective evaluations, total | 31,173 | **5,831** (5.3× fewer) |
| clean convergence (code 0) | 181 / 500 | **490 / 500** |

And the answers barely move. Comparing the optimum found under the loose setting
against the tight one:

- SSE: maximum absolute difference $1.2 \times 10^{-8}$, maximum relative
  difference $7.6 \times 10^{-9}$;
- $m$: maximum absolute difference $1.2 \times 10^{-5}$;
- $\omega$: maximum absolute difference $1.4 \times 10^{-5}$;
- the selected best $t_c$ is **2715 under both**.

Most striking: the loose setting produced a *worse* objective in only 53 of 500
cases, meaning that in the other 447 it was equal or better. **The tight
tolerance is not buying accuracy.** It cannot, because it never converges — it
terminates on line-search failure, and where that failure lands is essentially
arbitrary.

So `factr = 1e-08` costs roughly a factor of five in objective evaluations and
returns nothing for it. Correcting it requires no new dependency, no
restructuring, and no compiled code — only a changed default.

The one real cost is that estimates would move in their fifth decimal. That is
immaterial for interpretation, where $m$ and $\omega$ are read to two or three
decimals, but it does mean cached fits would no longer reproduce byte-identically
against the current ones.

## The combined measurement, done properly

Redone on a **fixed evaluation budget** — 500 candidates × 60 objective
evaluations each — so that every variant provably performs identical work:

| configuration | time | speedup |
|---|---:|---:|
| serial, symengine (as shipped) | 28.38 s | 1.00× |
| serial, R `crossprod` | 4.23 s | 6.70× |
| serial, Rcpp fused | 2.98 s | 9.52× |
| 8 cores, R `crossprod` | 0.86 s | 33.05× |
| 8 cores, Rcpp fused | **0.63 s** | **44.82×** |

These serial factors (6.7×, 9.5×) are lower than the component factors (10.6×,
22×) because the fixed budget also calls `eval_lppls()` on every evaluation, and
that cost is common to all variants — Amdahl again, one level down.

# Does symengine buy anything at all?

The investigation started from the assumption that symengine was costing
performance. Having measured it, the fuller question is whether it is buying
anything to offset that cost. Taking the possibilities in turn:

**Precision: no --- it is the worst of the six.** Section "Precision"
settles this against a constructed ground truth. symengine is the least accurate
variant at every conditioning level tested, by up to six orders of magnitude, and
its larger parameter error shows up as a genuinely worse residual.

**A more stable solve: no.** One might expect a symbolic solution to be exact
where a numeric one is not. It is not, because the expression is evaluated in the
same double precision as everything else, and a closed-form $4 \times 4$ solve
involves differences of products that cancel. Variants A and C confirm this
empirically: solving the identical matrix symbolically or with `solve()` gives
errors that agree to within a few percent.

**Speed: no.** It is 10.6× slower than the best pure-R alternative and 22× slower
than the Rcpp version.

**Avoiding a BLAS dependency: no.** R links a BLAS regardless; `crossprod()`
introduces no new requirement.

**Reusable symbolic machinery: not as used.** Symbolic differentiation could in
principle generate the gradient and Hessian expressions that `compute_X_matrix()`
and `compute_H_matrix()` currently hard-code by hand — which would be a real
argument for keeping a CAS in the package. But that is not what the code does:
symengine is used for exactly one $4 \times 4$ linear solve, and the derivative
matrices are hand-transcribed from the paper. (Hand-transcribed with an error, as
it happens: the asymmetric $H$ bug. A symbolically generated $H$ would have been
symmetric by construction, so the *unrealised* version of this argument is the
strongest case for a CAS anywhere in the package.)

**Documentation value: marginal, and cheaper by other means.** The symbolic setup
does record what is being solved. A four-line comment does the same without a
47.5 MB dependency.

So on the measurements alone symengine earns nothing as a *solver*. It is slower,
less accurate, and used for a single task that `solve()` performs better.

## Why it is kept anyway

There is a reason the benchmarks cannot see. The `"symengine"` method is the
calibration the thesis ran, and the symbolic solve together with the `sum_*`
helpers corresponds directly to the derivations in the thesis appendix. Keeping
it lets a reader reproduce the thesis numbers exactly and compare the code
against the equations it came from, rather than taking on trust that a faster
reimplementation agrees.

So `"symengine"` ships, unchanged, as a non-default option. It is not there
because it computes anything better; it is there because it computes the same
thing *in the form the thesis derived*.

# The CAS as a test oracle

The one place a computer-algebra system would genuinely pay --- deriving the MPL
gradient and Hessian rather than transcribing them --- is not where symengine was
being applied. That gap is now closed, though not as a runtime path.

`compute_X_matrix()` and `compute_H_matrix()` transcribe Filimonov (2017)
eq. (36) and (37) by hand. A CAS can derive them instead: state the model once,
and differentiate.

```{r cas-derive, eval=RECOMPUTE}
tau <- symengine::Symbol("tau"); m <- symengine::Symbol("m")
w   <- symengine::Symbol("w");   A <- symengine::Symbol("A")
B   <- symengine::Symbol("B");   C1 <- symengine::Symbol("C1")
C2  <- symengine::Symbol("C2")
f <- A + tau^m * (B + C1 * cos(w * log(tau)) + C2 * sin(w * log(tau)))
symengine::D(f, m)                    # a column of X
symengine::D(symengine::D(f, m), w)   # an entry of H
```

The derived expressions reproduce the hand-written matrices to
$3.6 \times 10^{-15}$, and confirm that the entries the hand-written code leaves
at zero --- every second derivative among the linear parameters --- are
identically zero. The `## The rest is zero` comment was therefore correct; the
defect was only the missing mirroring of `H[2,5]` and `H[2,6]`.

## Why not as a runtime path

The obvious next thought is to replace the hand-written loops with compiled
symbolic ones. Measured, that is a poor trade:

| X matrix (n = 2428) | median µs | vs hand-written |
|---|---:|---:|
| hand-written R loop | 510.9 | 1.0× |
| CAS (`DoubleVisitor`) | 868.3 | **0.6× --- slower** |
| Rcpp | 36.6 | 14.0× |

The CAS path is **1.7 times slower than what is already there**, because each
column is a separate visitor call over the whole vector, re-evaluating $\tau^m$
and the trigonometry per column instead of sharing them. Derivation itself is
free (under 0.01 s, once per session).

## Where it does pay

Its value is entirely at derivation time, which makes it a **test oracle** rather
than a runtime option. `tests/testthat/test-beta_method.R` derives the gradient
and Hessian symbolically, compares them against the hand-written implementations,
and asserts that $H$ is symmetric --- mixed partials commute, so
`D(D(f, a), b)` and `D(D(f, b), a)` are the same expression and an asymmetric $H$
cannot be produced.

That is precisely the check that would have caught the asymmetric-$H$ defect
automatically, and it will catch any future transcription slip in the same place,
at no runtime cost. The hand-written matrices remain the implementation, for the
same reason the `"symengine"` solver is kept: they are what the thesis derived,
and what a reader can compare against the appendix.

# The dependency question

An earlier draft of this document recommended dropping symengine altogether.
That recommendation was **not** adopted, and the reasoning is worth recording
because the measurements alone would have supported it.

The coupling is unusually clean, so removal would have been easy:

- symengine appears **only** in `R/helpers.R`, inside
  `create_beta_calculator()`;
- `lppls_f()`, `lppls_g()`, `lppls_h()` and all twelve `sum_*` helpers have
  **no callers anywhere else in the package** — they exist solely to feed it;
- the installed footprint is **47.5 MB**, and symengine requires a C++ toolchain
  to build from source.

What the measurements could not weigh is the reproduction value described in
section "Why it is kept anyway": the symengine path *is* the thesis
calibration, and it is the version a reader can check against the appendix. Since
it is now one option among four rather than the only implementation, its cost is
paid only by users who ask for it, while its benefit — an exactly reproducible
thesis calibration, and a second independent implementation to cross-check the
default against — is available to everyone. The `sum_*` helpers stay live for the
same reason.

The dependency also acquired a second job it is genuinely good at: the test
oracle of section "The CAS as a test oracle". So symengine now earns its
place twice over, in neither case as a solver.

One irony is worth stating plainly: adopting Rcpp would introduce a compilation
requirement that dropping symengine would have removed. Since symengine is
staying, that argument no longer cuts against Rcpp — the toolchain requirement is
already there.

# Recommendations and status

## Done

**The solver is now selectable, defaulting to `"crossprod"`.**
`create_beta_calculator(method =)` and `fit_lppls(..., beta_method =)` offer
`"crossprod"`, `"qr"`, `"chol"` and `"symengine"`. The default was chosen for
speed: within the admissible $0.1 \le m \le 0.9$ both `"crossprod"` and `"qr"`
are exact to roughly $10^{-12}$ or better, so the 1.3× speed difference decides
it, and `"qr"` remains one argument away for anyone widening the bounds.

Verified on the S&P 500 calibration that motivated the investigation:

| `beta_method` | time | $t_c$ | MPLE | LI(0.05) |
|---|---:|---:|---:|---|
| `"symengine"` | 87.5 s | 2717.00880445 | 2717 | [2696, 2757] |
| `"crossprod"` | **18.7 s** | 2717.00880445 | 2717 | [2696, 2757] |

**4.7× faster, with every reported quantity identical.**

**The CAS became a test oracle** rather than a runtime path, for the reasons in
section "The CAS as a test oracle".

**A related waste was removed.** `compute_mpl_loglik()` was calling
`create_beta_calculator()` **once per candidate $t_c$**, rebuilding the symbolic
solve 500 times in an SPX-scale fit at 1.73 ms each --- about 0.9 s per fit spent
constructing a function it then used once. The calculator is now built once and
threaded through, which also makes `beta_method` reach the MPL path.

## Still open

**Fix `factr`.** Roughly 5× fewer objective evaluations, no structural change, no
dependency change, and the current setting is not defensible on its own terms ---
it requests a tolerance below machine epsilon and guarantees line-search failure.
Changing the default from `1e-08` to `1e7` is a one-line change. Estimates move
in the fifth decimal, so cached fits would need regenerating; a direct comparison
found every reported thesis quantity unchanged and every figure visually
identical.

This is now the **largest remaining gain**, and it compounds with the solver
change rather than overlapping it: one cuts the number of objective evaluations,
the other the cost of each. Together they would take the SPX calibration from
about 84 seconds to the low single digits.

**Consider Rcpp for `compute_X_matrix()` and `compute_H_matrix()`.** 13.9× and
33.4× respectively, and unlike the solver these are explicit R loops with no
vectorised alternative --- the strongest case for compiled code in the package. It
introduces a build toolchain requirement, so it deserves a deliberate decision;
note that keeping symengine means that requirement already exists.

## Declined

**Internal parallelism.** The measured 4.8× at eight cores is real, but the axis
is already exploited at the experiment level, the restructuring is invasive,
reproducibility under random restarts is genuinely at risk, and nested
parallelism has already cost this project hours. A planned sweep layer over many
series would widen the outer axis further, making internal parallelism actively
counterproductive: cores spent inside `fit_lppls()` are cores unavailable to the
sweep.

# Reproducing this

Knit with `params$recompute = TRUE` to re-run everything (roughly 30 minutes, and
the numbers will differ by machine):

```r
rmarkdown::render("misc/bench_beta.Rmd", params = list(recompute = TRUE))
```

The benchmark harnesses are inlined in the chunks above. The Rcpp prototypes are
in `misc/` alongside this file; they are prototypes for measurement and are not
part of the package.

```{r bench-live, eval=RECOMPUTE}
## The component benchmark of section "The variants". The remaining harnesses --
## the profile, the end-to-end substitution, the parallel scaling and the factr
## comparison -- are described in full in their sections and are quick to
## reconstruct from them.
if (!requireNamespace("microbenchmark", quietly = TRUE)) {
  stop("Package 'microbenchmark' is needed to re-run these benchmarks.")
}
bc  <- lpplsF:::create_beta_calculator()
env <- environment(bc)

vals_from_basis <- function(b, log_p, nn) {
  f <- b$f; g <- b$g; h <- b$h
  sf <- sum(f); sg <- sum(g); sh <- sum(h)
  sff <- sum(f * f); sgg <- sum(g * g); shh <- sum(h * h)
  sfg <- sum(f * g); sfh <- sum(f * h); sgh <- sum(g * h)
  list(xx11 = nn, xx12 = sf, xx13 = sg, xx14 = sh,
       xx21 = sf, xx22 = sff, xx23 = sfg, xx24 = sfh,
       xx31 = sg, xx32 = sfg, xx33 = sgg, xx34 = sgh,
       xx41 = sh, xx42 = sfh, xx43 = sgh, xx44 = shh,
       xy1 = sum(log_p), xy2 = sum(log_p * f),
       xy3 = sum(log_p * g), xy4 = sum(log_p * h))
}

variants <- list(
  "A shipped (symengine)" = function(y, t, tc, m, w) bc(y, t, tc, m, w),
  "B shared + symengine" = function(y, t, tc, m, w) {
    v <- vals_from_basis(basis(t, tc, m, w), y, length(t))
    c(A = drop(do.call(env$A_func, v)),  B  = drop(do.call(env$B_func, v)),
      C1 = drop(do.call(env$C1_func, v)), C2 = drop(do.call(env$C2_func, v)))
  },
  "C shared + solve" = function(y, t, tc, m, w) {
    v <- vals_from_basis(basis(t, tc, m, w), y, length(t))
    XX <- matrix(c(v$xx11, v$xx12, v$xx13, v$xx14, v$xx21, v$xx22, v$xx23, v$xx24,
                   v$xx31, v$xx32, v$xx33, v$xx34, v$xx41, v$xx42, v$xx43, v$xx44),
                 4L, 4L, byrow = TRUE)
    s <- solve(XX, c(v$xy1, v$xy2, v$xy3, v$xy4))
    c(A = s[1], B = s[2], C1 = s[3], C2 = s[4])
  },
  "D crossprod + solve" = beta_crossprod,
  "E .lm.fit (QR)" = function(y, t, tc, m, w) {
    b <- basis(t, tc, m, w)
    s <- .lm.fit(cbind(1, b$f, b$g, b$h), y)$coefficients
    c(A = s[1], B = s[2], C1 = s[3], C2 = s[4])
  },
  "F chol" = function(y, t, tc, m, w) {
    b <- basis(t, tc, m, w); X <- cbind(1, b$f, b$g, b$h)
    s <- chol2inv(chol(crossprod(X))) %*% crossprod(X, y)
    c(A = s[1], B = s[2], C1 = s[3], C2 = s[4])
  })

set.seed(1)
n  <- 2428; tt <- seq_len(n); tc <- n + 21
y  <- eval_lppls(tt, 8, -0.025, 0.0015, 0.001, tc, 0.6, 8) + rnorm(n) * 0.01
ref <- variants[[1]](y, tt, tc, 0.6, 8)
acc <- vapply(variants, function(f) max(abs((f(y, tt, tc, 0.6, 8) - ref) / ref)), numeric(1))
mb  <- do.call(microbenchmark::microbenchmark,
               c(lapply(variants, function(f) bquote(.(f)(y, tt, tc, 0.6, 8))),
                 list(times = 200L)))
knitr::kable(
  data.frame(variant = names(variants),
             `max rel. deviation` = sprintf("%.2e", acc),
             `median us` = round(tapply(mb$time, mb$expr, median) / 1000, 1),
             check.names = FALSE),
  row.names = FALSE)
```
