---
title: "lpplsF equivalence checks: thesis implementation vs. package"
author: "mhv"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

Before comparing the *downstream* Lagrange regularization, this file verifies the
layers underneath it, in order of decreasing determinism:

* **Stage A** — model evaluation: `eval_lppls()` vs. `thesis_fns$LPPLS()`.
* **Stage B** — OLS / objective: each side vs. an independent normal-equation reference.
* **Stage C** — optimizer at a single fixed start (no random restarts).
* **Stage D** — full fitter incl. MPL likelihood intervals.

A discrepancy at any stage invalidates the "shared frozen fits" assumption that the
Lagrange comparison rests on. Stages A and B are deterministic and exact; Stage C is
deterministic only if the init-search does not overwrite the supplied start; Stage D
may only be reproducible up to tolerance.

# Setup

```{r message = FALSE, warning = FALSE}
find_pkg_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = FALSE)
  while (!file.exists(file.path(path, "DESCRIPTION"))) {
    parent <- dirname(path)
    if (parent == path) stop("DESCRIPTION not found above ", start)
    path <- parent
  }
  path
}
root <- find_pkg_root()

## Corrected package (exports eval_lppls, SSE, fit_lppls; internal create_beta_calculator)
devtools::load_all(root)

## Buggy thesis reference, isolated in its own environment
thesis_fns <- new.env()
source(file.path(root, "misc", "lpplsF_thesis_functions.R"), local = thesis_fns)

## Package OLS calculator (internal -> namespace-qualified)
pkg_beta <- lpplsF:::create_beta_calculator()
```

## Reporting helper

```{r}
report <- function(label, a, b, tol = 1e-10) {
  a <- unname(as.numeric(a)); b <- unname(as.numeric(b))
  dd <- abs(a - b)
  d  <- if (length(dd) == 0 || all(is.na(dd))) NA_real_ else max(dd, na.rm = TRUE)
  ok <- isTRUE(d <= tol)
  cat(sprintf("%-42s max|diff| = %-11s %s\n", label,
              formatC(d, format = "e", digits = 3), if (ok) "OK" else "**CHECK**"))
  invisible(ok)
}
```

## Independent reference (first-principles OLS + SSE over t < tc)

```{r}
ref_fit <- function(log_p, t, tc, m, omega) {
  keep <- t < tc
  tt <- t[keep]; y <- log_p[keep]
  tau <- (tc - tt)^m
  d   <- omega * log(tc - tt)
  X   <- cbind(1, tau, tau * cos(d), tau * sin(d))
  beta <- drop(solve(crossprod(X), crossprod(X, y)))
  fitted <- as.vector(X %*% beta)
  list(beta = c(A = beta[1], B = beta[2], C1 = beta[3], C2 = beta[4]),
       sse  = sum((y - fitted)^2))
}
```

## Shared e014 data and window

```{r}
sigma <- 0.05; N <- 2000; tc <- N + 1
truth <- list(A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001, tc = tc, m = 0.6, omega = 8)

## Reproduce the e014 draw pattern exactly: seed once, then N single draws.
set.seed(3630)
sim <- numeric(N)
for (i in 1:N) {
  sim[i] <- thesis_fns$LPPLS(i, truth$A, truth$B, truth$C1, truth$C2,
                             truth$tc, truth$m, truth$omega) + rnorm(1) * sigma
}

hold_out <- 100; fh <- 150
time_ID  <- 1:N
log_pr   <- sim
t_mod    <- head(time_ID, N - hold_out)        # [T1, T2] modelling window
y_mod    <- head(log_pr,  N - hold_out)

## Fit settings shared by both fitters (as in e014)
lower <- c(0.1, 6, -0.03, 0.8); upper <- c(0.9, 13, -1e-14, 1e6)
tc_init <- length(t_mod) + (fh / 2)            # = N - hold_out + fh/2 in window-local terms
m_init  <- 0.2; o_init <- 12; num_searches <- 1
```

# Stage A — model evaluation (deterministic, exact)

Both functions are vectorized over `t` and share the same argument order, so compare
directly across a grid of nonlinear parameters and all four modes.

```{r}
grid <- expand.grid(tc = c(2001, 2100), m = c(0.3, 0.6, 0.9), omega = c(6, 8, 12))
tA   <- 1:1995

for (mode in 0:3) {
  worst <- 0
  for (r in seq_len(nrow(grid))) {
    g <- grid[r, ]
    ## mode 2/3 use T1/T2 scale params; pass them explicitly and identically.
    pkg <- eval_lppls(tA, truth$A, truth$B, truth$C1, truth$C2, g$tc, g$m, g$omega,
                      mode = mode, T1 = 500, T2 = 1990, omega2 = 0.5, omega3 = 0.2)
    thx <- thesis_fns$LPPLS(tA, truth$A, truth$B, truth$C1, truth$C2, g$tc, g$m, g$omega,
                            mode = mode, T1 = 500, T2 = 1990, omega2 = 0.5, omega3 = 0.2)
    worst <- max(worst, max(abs(pkg - thx), na.rm = TRUE))
  }
  report(sprintf("Stage A: eval_lppls vs LPPLS (mode %d)", mode), worst, 0)
}
```

Mode 0 is the only mode used by e014/e016b and must match to floating point. Modes 1–3
are not exercised by the reproduction. **Expected divergence:** once the `omega * log(tau)`
fix is applied to `eval_lppls()` mode 3 (the `+` you confirmed should be `*`), mode 3 will
report a mismatch here, because the thesis `LPPLS` still carries the original `+`. That
mismatch is the correction, not a regression — leave the thesis side untouched.

# Stage B — OLS / objective vs. independent reference (deterministic, exact)

## B1: package OLS calculator vs. reference, on a grid

```{r}
ok <- TRUE
for (r in seq_len(nrow(grid))) {
  g  <- grid[r, ]
  rf <- ref_fit(y_mod, t_mod, g$tc, g$m, g$omega)
  pb <- pkg_beta(y_mod, t_mod, g$tc, g$m, g$omega)
  ok <- report(sprintf("Stage B1: pkg beta (tc=%g,m=%g,w=%g)", g$tc, g$m, g$omega),
               c(pb["A"], pb["B"], pb["C1"], pb["C2"]), rf$beta) && ok
}
```

The thesis `beta_calculator` is a closure inside `lpplsF_thesis()` and is not reachable,
so it is validated indirectly in B2 at the point the thesis fitter actually returns.

## B2: reference SSE matches the reference OLS round-trip

```{r}
g  <- grid[1, ]
rf <- ref_fit(y_mod, t_mod, g$tc, g$m, g$omega)
keep <- t_mod < g$tc
fitted <- eval_lppls(t_mod[keep], rf$beta["A"], rf$beta["B"], rf$beta["C1"], rf$beta["C2"],
                     g$tc, g$m, g$omega, mode = 0)
report("Stage B2: ref SSE vs eval_lppls round-trip",
       rf$sse, sum((y_mod[keep] - fitted)^2))
```


Constant? Or structured?
```{r}
plot(th$mpl_output$LL, type = "l"); lines(pk$mpl_output$LL, col = "red")
summary(th$mpl_output$LL - pk$mpl_output$LL)   ## constant? or structured?
```


# Stage C — optimizer at a single fixed start (F2, no MPL)

`num_searches = 1` means the supplied init is used with no random restarts; the only
nondeterminism is the init-search that overwrites `m_init`/`o_init` when they do not yield
`B < 0`. Run each fitter twice under the same seed to confirm self-reproducibility, then
compare across implementations.

```{r}
get_coef <- function(f1) {
  pick <- function(nm, pos) if (!is.null(f1[[nm]])) unname(f1[[nm]]) else unname(f1[[pos]])
  c(tc = pick("tc", 1), m = pick("m", 2), omega = pick("omega", 3),
    A = pick("A", 4), B = pick("B", 5), C1 = pick("C1", 6), C2 = pick("C2", 7),
    value = if (!is.null(f1[["value"]])) unname(f1[["value"]]) else unname(f1[[9]]))
}

run_thesis <- function(seed, mode = "F2") {
  set.seed(seed)
  thesis_fns$lpplsF_thesis(
    time_ID = time_ID, log_price = log_pr, fh = fh, hold_out = hold_out,
    lower = lower, upper = upper, tc_init = tc_init, m_init = m_init, o_init = o_init,
    num_searches = num_searches, mode = mode, mpl_cutoff = c(0.05, 0.1, 0.5),
    mpl_plot = 0, fp = 0, cp = 0, sp = 0, tp = c(0,0,0), tp_id = 1, pp = 0, mp = 0,
    factr = 1e-08, fb = 0)
}
run_pkg <- function(seed, mode = "F2") {
  set.seed(seed)
  fit_lppls(
    log_price = log_pr, fh = fh, hold_out = hold_out,
    lower = lower, upper = upper, tc_init = tc_init, m_init = m_init, o_init = o_init,
    num_searches = num_searches, mode = mode, mpl_cutoff = c(0.05, 0.1, 0.5),
    mpl_plot = FALSE, fp = FALSE, cp = FALSE, sp = FALSE, tp = c(0,0,0), tp_id = 1,
    pp = FALSE, mp = FALSE, factr = 1e-08, fb = FALSE)
}

## lpplsF_thesis() only completes in MPL mode (mpl_output is assigned solely in
## the MPL branch). MPL does not alter fit[[1]], so run both in MPL and compare
## only fit[[1]] -- the optimizer core. The LI machinery is compared in Stage D.

th_fit <- run_thesis(1, "MPL");  th_a <- get_coef(th_fit$fit[[1]])
pk_fit <- run_pkg(1,    "MPL");  pk_a <- get_coef(pk_fit$fit[[1]])
th_b <- get_coef(run_thesis(1, "MPL")$fit[[1]])   # self-reproducibility
pk_b <- get_coef(run_pkg(1,    "MPL")$fit[[1]])

report("Stage C: thesis self-reproducible",  th_a, th_b)
report("Stage C: package self-reproducible", pk_a, pk_b)
report("Stage C: thesis vs package fit[[1]]", th_a, pk_a, tol = 1e-6)
```

If the cross-implementation line mismatches while both self-reproduce, Stage A + B have
already cleared the model and OLS math, so the gap is in the optimizer wrapper: bounds
handling, `optim` control, or the init-search overwriting the start differently. Re-run
the two fits with `fb = TRUE` (uncomment below) to see whether `m_init`/`o_init` were
overwritten on either side.

```{r eval = FALSE}
invisible(run_thesis(1, "F2"))   # watch for init overwrite messages
invisible(run_pkg(1, "F2"))
```

# Stage D — full fitter incl. MPL (may only match up to tolerance)

Compares the MPL point estimate and likelihood-interval bounds at each cutoff.

```{r}
th <- th_fit
pk <- pk_fit

report("Stage D: MPL fit (tc,m,omega,A,B,C1,C2,value)",
       get_coef(th$fit[[1]]), get_coef(pk$fit[[1]]), tol = 1e-6)

## tc_hat_mpl and the three LI bands. Wrapped defensively in case the
## mpl_output structures differ between implementations.
try({
  report("Stage D: tc_hat_mpl",
         th$mpl_output$tc_hat_mpl, pk$mpl_output$tc_hat_mpl, tol = 1e-6)
  for (j in 1:3) {
    report(sprintf("Stage D: LI[[%d]] (lower, upper)", j),
           th$mpl_output$LI[[j]], pk$mpl_output$LI[[j]], tol = 1e-6)
  }
})
```



# Summary

Read top to bottom: the first stage that mismatches localizes the discrepancy. A through
C should be exact (modulo the intended mode-3 correction); only D is allowed to fall back
to tolerance. Once every stage here is clean, the fitted vectors are a verified shared
input, and the Lagrange knob harness can be run on top of them with confidence that any
remaining difference is genuinely downstream.
```






