---
title: "Performance of `lpplsF`"
subtitle: "What was changed, why it was slow, and what each change bought"
author: "lpplsF"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: kable
params:
  live:
    label: "Run the benchmarks on this machine (FALSE = show the recorded numbers only)"
    value: true
  scale:
    label: "Workload for the live whole-fit comparison"
    value: "quick"
    choices: ["quick", "spx"]
  reps:
    label: "Repetitions for the live whole-fit timings"
    value: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

pkg_root <- "."
for (up in c(".", "..", file.path("..", ".."))) {
  if (file.exists(file.path(up, "DESCRIPTION"))) { pkg_root <- up; break }
}
suppressMessages(pkgload::load_all(pkg_root, quiet = TRUE))

LIVE <- isTRUE(params$live)
REPS <- as.integer(params$reps)

## Timing helper, in microseconds per call. The repetition count is calibrated
## rather than fixed: system.time() has millisecond resolution, so a fixed count
## on a ~50 us call measures the clock, not the code. Keep doubling until the
## accumulated time is long enough for the quantisation to be irrelevant.
us_per_call <- function(expr, min_time = 0.4, max_reps = 5e5, env = parent.frame()) {
  e <- substitute(expr)
  invisible(eval(e, env))                                   # warm
  reps <- 2L
  repeat {
    el <- system.time(for (i in seq_len(reps)) eval(e, env))[["elapsed"]]
    if (el >= min_time || reps >= max_reps) break
    reps <- min(max_reps, max(reps * 2, ceiling(reps * 1.3 * min_time / max(el, 1e-3))))
  }
  1e6 * el / reps
}
sec_median <- function(expr, reps = REPS, env = parent.frame()) {
  e <- substitute(expr)
  ts <- numeric(reps)
  for (i in seq_len(reps)) { gc(FALSE); ts[i] <- system.time(eval(e, env))[["elapsed"]] }
  stats::median(ts)
}
fmt <- function(x, d = 1) formatC(x, format = "f", digits = d, big.mark = ",")
speedup <- function(before, after) sprintf("%.1fx", before / after)
```

```{r data}
spx <- read.csv(system.file("extdata", "SPX.csv", package = "lpplsF"))
names(spx)[1] <- "Date"
spx$logpris <- log(spx$spindx)
df_spx  <- spx[11800:14727, ]
n_spx   <- nrow(df_spx) - 500          # 2428 calibration points
y_spx   <- df_spx$logpris[seq_len(n_spx)]
t_spx   <- seq_len(n_spx)
tc_spx  <- 2717; m_spx <- 0.9; w_spx <- 6.1597
Psi_spx <- c(m_spx, w_spx, 8.11798574, -9.089053e-4, 7.66909749e-5, 3.88299997e-5)

## A smaller synthetic series for the live whole-fit comparison.
set.seed(3630)
n_q  <- 1000L
y_q  <- eval_lppls(seq_len(n_q), 8, -0.025, 0.0015, 0.001, n_q + 61, 0.6, 8) +
  rnorm(n_q) * 0.02
```

# What this document is

`lpplsF`'s calibration was made about **27 times faster** over a sequence of
changes, none of which altered a reported quantity. This report accounts for each
change: what it was, why the original was slow, and what it bought.

It is a **hybrid** of recorded and live measurement, for a reason worth stating
plainly. Measuring a speed-up needs both arms, and most of the "before" arms no
longer exist in the package — the looped MPL matrices, the `tibble`/`dplyr`
bookkeeping, the two-pass objective and the old trace sizing were all removed.
Only two survive as user-facing parameters: `beta_method = "symengine"` and
`factr = 1e-08`.

So:

* **§2's arc is recorded**, from the sessions in which each change was made. It
  cannot be re-derived here, because the code it measures is in git history.
* **§3 onward is live** wherever the comparison is reachable — either through a
  parameter, or by reproducing the previous implementation inline as a clearly
  labelled reference function. Those reference functions are in this document,
  not in the package.

Set `params$live = FALSE` to render the recorded numbers alone.

```{r env}
data.frame(
  item = c("R", "platform", "BLAS", "lpplsF", "live benchmarks", "workload"),
  value = c(R.version.string,
            R.version$platform,
            basename(extSoftVersion()[["BLAS"]]),
            as.character(utils::packageVersion("lpplsF")),
            if (LIVE) "yes" else "no (recorded numbers only)",
            params$scale)
)
```

# The arc (recorded)

The reference workload throughout is the S&P 500 calibration: $n = 2428$
calibration points, `fh = 500` candidate $t_c$, `num_searches = 1`, `mode = "MPL"`.
Each row adds one change to the row above it.

```{r arc}
arc <- data.frame(
  stage = c("Original (symengine solver, factr = 1e-08)",
            "+ crossprod solver (`beta_method`)",
            "+ factr = 1e7 (optim's own default)",
            "+ vectorised MPL matrices",
            "+ flattened per-tc bookkeeping",
            "+ fused objective"),
  spx_mpl_s = c(87.5, 18.7, 8.63, 5.09, 4.45, 3.26),
  stringsAsFactors = FALSE
)
arc$cumulative <- sprintf("%.1fx", arc$spx_mpl_s[1] / arc$spx_mpl_s)
arc$step <- c("--", sprintf("%.2fx", head(arc$spx_mpl_s, -1) / tail(arc$spx_mpl_s, -1)))
names(arc) <- c("stage", "SPX MPL fit (s)", "cumulative", "this step")
arc
```

Two provenance notes. The first two rows are quoted from the session that made
those changes; rows 3-6 were measured together under one protocol (median of
three, each in a fresh R process). The 8.63 s in row 3 is that protocol's
measurement of the state the earlier session recorded as 8.3 s — the difference
is measurement conditions, not code.

**Nothing in the arc moved a reported quantity.** Across every stage the SPX fit
returns $t_c = 2717.0088\ldots$, $m = 0.9$, $\omega = 6.16$, MPLE $= 2717$ and
likelihood intervals $[2696, 2757]$, $[2696, 2757]$, $[2696, 2744]$. The
estimates shift in the 11th-12th significant digit where a change altered the
arithmetic; §8 says which changes did that and which were bit-identical.

# The linear solver

At fixed $(t_c, m, \omega)$ the model is linear in $(A, B, C_1, C_2)$, so each
objective evaluation solves a $4 \times 4$ least-squares problem. The thesis
implementation solved it symbolically with **symengine**, building the entries of
$X'X$ through `sum_*` helpers that recomputed the $\tau^m$/$\cos$/$\sin$ basis
for *every entry* — 16 recomputations of the same three vectors per call.

The fix was to build the basis once and hand $X$ to a numerical solver.
`beta_method` now selects among four, with the symbolic one retained so the
thesis calibration stays reproducible.

```{r solvers, eval=LIVE}
sol <- data.frame(method = c("crossprod", "qr", "chol", "symengine"),
                  stringsAsFactors = FALSE)
sol$us_per_call <- vapply(sol$method, function(mm) {
  f <- create_beta_calculator(mm)
  us_per_call(f(y_spx, t_spx, tc_spx, m_spx, w_spx))
}, numeric(1))
sol$relative <- sprintf("%.1fx", sol$us_per_call / min(sol$us_per_call))
names(sol) <- c("beta_method", "us per call (n = 2428)", "vs fastest")
sol
```

```{r solvers-note, results="asis", eval=LIVE}
cat("The ordering is the point, not the absolute numbers: the symbolic solver is",
    "an order of magnitude behind, and the cost is in forming `X'X`, not in the",
    "symbolic step itself.\n")
```

# The convergence tolerance

`fit_lppls()` inherited `factr = 1e-08` from the thesis. `factr` **multiplies
machine epsilon**, so that asks for a relative tolerance of
$10^{-8} \times 2.2\times10^{-16} \approx 2\times10^{-24}$ — unsatisfiable in
double precision. L-BFGS-B therefore never converged; it ran until its line
search failed and returned convergence code 52, having burned evaluations to no
purpose. The default is now `1e7`, which is `optim()`'s own.

```{r factr, eval=LIVE}
bc <- create_beta_calculator("crossprod")
sse2 <- function(par, tc, log_p, t) {
  b <- bc(log_p, t, tc, par[1], par[2])
  sum((log_p - eval_lppls(t, b["A"], b["B"], b["C1"], b["C2"],
                          tc, par[1], par[2], mode = 0))^2, na.rm = TRUE)
}
## Aggregate over a stretch of candidate tc: whether the tolerance bites at all
## depends on the individual sub-problem, so a single tc proves nothing.
tc_scan <- n_spx + seq(1, 500, by = 5)          # 100 candidates across the range
fr <- do.call(rbind, lapply(c(1e-08, 1e7), function(fc) {
  o <- lapply(tc_scan, function(tc)
    stats::optim(c(0.2, 12), sse2, tc = tc, log_p = y_spx, t = t_spx,
                 lower = c(0.1, 6), upper = c(0.9, 13),
                 method = "L-BFGS-B", control = list(factr = fc)))
  data.frame(factr = format(fc),
             evaluations = sum(vapply(o, function(z) z$counts[["function"]], 0)),
             clean_convergence = sprintf("%d / %d",
               sum(vapply(o, function(z) z$convergence, 0) == 0), length(o)),
             worst_objective = sprintf("%.10f", max(vapply(o, function(z) z$value, 0))),
             stringsAsFactors = FALSE)
}))
fr$evaluations_ratio <- sprintf("%.1fx", fr$evaluations / min(fr$evaluations))
names(fr) <- c("factr", "objective evaluations (100 tc)", "clean convergence",
               "worst optimum found", "cost vs factr = 1e7")
fr
```

Both tolerances find the same optima; one of them stops when it gets there. Where
`factr = 1e-08` does not return code 0, the optimiser ended in line-search
failure rather than convergence — it could not satisfy a tolerance below what
double precision represents.

# Hoisting the solver out of the MPL loop

`compute_mpl_loglik()` called `create_beta_calculator()` **once per candidate**
$t_c$. For the symbolic method that means rebuilding four compiled visitor
closures 500 times per fit. The calculator is now built once by `fit_lppls()` and
passed down.

```{r hoist, eval=LIVE}
h <- data.frame(
  what = c("create_beta_calculator('symengine')", "create_beta_calculator('crossprod')"),
  us_per_call = c(us_per_call(create_beta_calculator("symengine")),
                  us_per_call(create_beta_calculator("crossprod"))))
h$cost_per_fit_s <- 500 * h$us_per_call / 1e6
names(h) <- c("constructing the calculator", "us per construction", "x500 per fit (s)")
h
```

# Vectorising the MPL matrices

`compute_X_matrix()` and `compute_H_matrix()` walked the series in an R `for`
loop, one observation at a time — $X$ twice and $H$ once per candidate $t_c$, so
1,500 loop traversals of length $n$ per MPL fit. Both are pure elementwise
arithmetic and vectorise directly.

This was the single largest remaining win, and it needed no compiler. The loops
below are the previous implementations, reproduced for comparison.

```{r mplmat, eval=LIVE}
compute_X_loop <- function(Psi, tc, t) {
  n <- length(t); X <- matrix(0, n, 6)
  m <- Psi[1]; omega <- Psi[2]; b <- Psi[4]; c1 <- Psi[5]; c2 <- Psi[6]
  for (i in seq_len(n)) {
    tau <- tc - t[i]; lt <- log(tau); tm <- tau^m
    ct <- cos(omega * lt); st <- sin(omega * lt)
    X[i, 1] <- tm * lt * (b + c1 * ct + c2 * st)
    X[i, 2] <- tm * lt * (-c1 * st + c2 * ct)
    X[i, 3] <- 1; X[i, 4] <- tm; X[i, 5] <- tm * ct; X[i, 6] <- tm * st
  }
  X
}
compute_H_loop <- function(Psi, tc, log_p, t) {
  n <- length(t); H <- matrix(0, 6, 6)
  m <- Psi[1]; omega <- Psi[2]; a <- Psi[3]
  b <- Psi[4]; c1 <- Psi[5]; c2 <- Psi[6]
  for (i in seq_len(n)) {
    tau <- tc - t[i]; lt <- log(tau); tm <- tau^m
    ct <- cos(omega * lt); st <- sin(omega * lt)
    res <- log_p[i] - (a + tm * (b + c1 * ct + c2 * st))
    H[1, 1] <- H[1, 1] + res * tm * lt^2 * (b + c1 * ct + c2 * st)
    H[1, 2] <- H[1, 2] + res * tm * lt^2 * (-c1 * st + c2 * ct)
    H[1, 4] <- H[1, 4] + res * tm * lt
    H[1, 5] <- H[1, 5] + res * tm * lt * ct
    H[1, 6] <- H[1, 6] + res * tm * lt * st
    H[2, 2] <- H[2, 2] + res * (-1) * tm * lt^2 * (c1 * ct + c2 * st)
    H[2, 5] <- H[2, 5] + res * (-1) * tm * lt * st
    H[2, 6] <- H[2, 6] + res * tm * lt * ct
  }
  H[2, 1] <- H[1, 2]; H[4, 1] <- H[1, 4]; H[5, 1] <- H[1, 5]
  H[6, 1] <- H[1, 6]; H[5, 2] <- H[2, 5]; H[6, 2] <- H[2, 6]
  H
}

mm <- data.frame(
  routine = c("compute_X_matrix", "compute_H_matrix"),
  loop_us = c(us_per_call(compute_X_loop(Psi_spx, tc_spx, t_spx)),
              us_per_call(compute_H_loop(Psi_spx, tc_spx, y_spx, t_spx))),
  vec_us  = c(us_per_call(lpplsF:::compute_X_matrix(Psi_spx, tc_spx, t_spx)),
              us_per_call(lpplsF:::compute_H_matrix(Psi_spx, tc_spx, y_spx, t_spx))),
  stringsAsFactors = FALSE)
mm$speedup <- sprintf("%.1fx", mm$loop_us / mm$vec_us)
mm$identical <- c(
  identical(compute_X_loop(Psi_spx, tc_spx, t_spx),
            lpplsF:::compute_X_matrix(Psi_spx, tc_spx, t_spx)),
  identical(compute_H_loop(Psi_spx, tc_spx, y_spx, t_spx),
            lpplsF:::compute_H_matrix(Psi_spx, tc_spx, y_spx, t_spx)))
mm$per_fit_saved_s <- (mm$loop_us - mm$vec_us) * c(1000, 500) / 1e6
names(mm) <- c("routine", "loop (us)", "vectorised (us)", "speed-up",
               "bit-identical", "saved per MPL fit (s)")
mm
```

The `bit-identical` column is the important one: the vectorised forms return
exactly the same matrices, so no figure or estimate moved.

A related one-word change sits next to them. `compute_mpl_loglik()` computed
$X'X$ as `crossprod(X, X)`, the two-argument form, which dispatches to a general
matrix-matrix product. Written `crossprod(X)` it takes the symmetric-rank-update
path instead: slightly cheaper, and exactly symmetric by construction — which
matters, since $\det(X'X - H)$ is the observed information only when both terms
are symmetric.

```{r syrk, eval=LIVE}
Xs <- lpplsF:::compute_X_matrix(Psi_spx, tc_spx, t_spx)
data.frame(
  form = c("crossprod(X, X)  [general product]", "crossprod(X)  [symmetric update]"),
  us_per_call = c(us_per_call(crossprod(Xs, Xs)), us_per_call(crossprod(Xs))),
  exactly_symmetric = c(identical(crossprod(Xs, Xs), t(crossprod(Xs, Xs))),
                        identical(crossprod(Xs), t(crossprod(Xs)))))
```

# Flattening the per-`tc` bookkeeping

Everything in the `tc` loop that was not the optimisation was `tibble` row
assignment, two `dplyr::arrange()` calls and a `dplyr::filter()` per candidate,
plus an `rbind()` that regrew the results table on every iteration — quadratic in
`fh`. It also built and sorted two tables per candidate that nothing ever read.

It is now preallocated vectors, assembled into the documented tibbles once after
the loop. The block below reproduces the old per-candidate work.

```{r bookkeeping, eval=LIVE}
tbl0 <- tibble::tibble(ID = 1L, value = 5.3, tc = 2717, m = 0.9, omega = 6.16,
                       A = 8.1, B = -9e-4, C1 = 7.7e-5, C2 = 3.9e-5, D = 1.5)
old_block <- function(fh = 500) {
  best <- tbl0[rep(1, fh), ]; acc <- tbl0[0, ]
  for (k in seq_len(fh)) {
    tmp <- tbl0
    tmp[1, ] <- list(1L, 5.3, 2429 + k, 0.9, 6.16, 8.1, -9e-4, 7.7e-5, 3.9e-5, 1.5)
    s1 <- dplyr::arrange(tmp, value)
    s2 <- dplyr::arrange(dplyr::filter(tmp, B < 0), value)
    best[k, ] <- s1[1, ]
    if (nrow(s2) > 0) acc <- rbind(acc, s2[1, ])
  }
  list(best, acc)
}
## The replacement, faithfully: the same ten columns are accumulated in
## preallocated vectors and assembled into the same two tibbles after the loop.
new_block <- function(fh = 500) {
  b <- replicate(10, numeric(fh), simplify = FALSE)
  f <- replicate(10, numeric(fh), simplify = FALSE)
  n_keep <- 0L
  for (k in seq_len(fh)) {
    s_val <- 5.3; s_B <- -9e-4
    row <- list(1, 5.3, 2429 + k, 0.9, 6.16, 8.1, -9e-4, 7.7e-5, 3.9e-5, 1.5)
    j <- which.min(s_val); if (!length(j)) j <- 1L
    for (c in 1:10) b[[c]][k] <- row[[c]]
    if (length(which(s_B < 0))) {
      n_keep <- n_keep + 1L
      for (c in 1:10) f[[c]][n_keep] <- row[[c]]
    }
  }
  kept <- seq_len(n_keep)
  list(tibble::tibble(ID = as.integer(b[[1]]), value = b[[2]], tc = b[[3]],
                      m = b[[4]], omega = b[[5]], A = b[[6]], B = b[[7]],
                      C1 = b[[8]], C2 = b[[9]], D = b[[10]]),
       tibble::tibble(ID = as.integer(f[[1]][kept]), value = f[[2]][kept],
                      tc = f[[3]][kept], m = f[[4]][kept], omega = f[[5]][kept],
                      A = f[[6]][kept], B = f[[7]][kept], C1 = f[[8]][kept],
                      C2 = f[[9]][kept], D = f[[10]][kept]))
}
bk <- data.frame(
  implementation = c("tibble + dplyr + growing rbind", "preallocated vectors"),
  seconds_per_fit = c(us_per_call(old_block(), min_time = 1.5) / 1e6,
                      us_per_call(new_block(), min_time = 1.5) / 1e6))
bk$per_tc_us <- 1e6 * bk$seconds_per_fit / 500
names(bk) <- c("implementation (500 candidate tc)", "s per fit", "us per tc")
bk
```

This is a fixed cost per fit that neither vectorising nor a compiler would have
touched — it is not numerics at all.

# The trace replay

The BFGS trace plots reconstruct the optimiser's path by re-running L-BFGS-B with
`maxit = 1, 2, 3, ...`. The number of steps was taken from `optim()`'s
`counts["function"]`, which counts **calls to the objective, not iterations** —
in practice 2 to 12 times the number of steps. The trace was therefore padded
with points sitting exactly on top of the converged one; under the old
`factr = 1e-08`, where the optimiser never converged, e008's trace held 91 points
for an 11-step path.

The length is now discovered: extend `maxit` until the iterate stops changing.

```{r trace, eval=LIVE}
steps_to_settle <- function(tc) {
  prev <- NULL
  for (k in 1:200) {
    p <- stats::optim(c(0.2, 12), sse2, tc = tc, log_p = y_spx, t = t_spx,
                      lower = c(0.1, 6), upper = c(0.9, 13),
                      method = "L-BFGS-B", control = list(maxit = k))$par
    if (!is.null(prev) && identical(p, prev)) return(k - 1L)
    prev <- p
  }
  NA_integer_
}
tr <- do.call(rbind, lapply(c(2429, 2600, 2717, 2928), function(tc) {
  o <- stats::optim(c(0.2, 12), sse2, tc = tc, log_p = y_spx, t = t_spx,
                    lower = c(0.1, 6), upper = c(0.9, 13),
                    method = "L-BFGS-B", control = list(factr = 1e7))
  data.frame(tc = tc, counts_function = o$counts[["function"]],
             actual_steps = steps_to_settle(tc))
}))
tr$over_run <- sprintf("%.1fx", tr$counts_function / tr$actual_steps)
names(tr) <- c("tc", "counts['function'] (old length)", "actual steps", "over-run")
tr
```

# Fusing the objective

Each objective evaluation built the $\tau^m$/$\cos$/$\sin$ basis to solve for
$\hat\beta$, discarded it, and then rebuilt the identical quantities inside
`eval_lppls()` to form the fitted values. The `"crossprod"` path now keeps the
design matrix and takes the fitted values as $X\hat\beta$ — one pass instead of
two.

`misc/fused_objective.md` is the full mathematical account, including why the two
forms are algebraically identical but differ by about one unit in the last place
of the SSE.

```{r fused, eval=LIVE}
sse_two_pass <- function(log_p, t, tc, m, omega) {
  b <- bc(log_p, t, tc, m, omega)
  sum((log_p - eval_lppls(t, b["A"], b["B"], b["C1"], b["C2"],
                          tc, m, omega, mode = 0))^2, na.rm = TRUE)
}
sse_fused <- lpplsF:::create_sse_calculator("crossprod")
fu <- data.frame(
  implementation = c("two-pass (solve, then re-evaluate the model)",
                     "fused (solve, then X %*% beta)"),
  us_per_call = c(us_per_call(sse_two_pass(y_spx, t_spx, tc_spx, m_spx, w_spx)),
                  us_per_call(sse_fused(y_spx, t_spx, tc_spx, m_spx, w_spx))))
fu$speedup <- c("--", sprintf("%.2fx", fu$us_per_call[1] / fu$us_per_call[2]))
a <- sse_two_pass(y_spx, t_spx, tc_spx, m_spx, w_spx)
b <- sse_fused(y_spx, t_spx, tc_spx, m_spx, w_spx)
fu$relative_difference <- c("--", sprintf("%.2e", abs(a - b) / b))
names(fu) <- c("implementation", "us per call", "speed-up", "rel. difference in SSE")
fu
```

Since the objective is evaluated **26,890 times** in a single SPX F2 calibration
— 500 candidate $t_c$, each driving a search whose finite-difference gradients
cost about 54 evaluations — a saving of tens of microseconds per call is seconds
per fit.

# Live whole-fit comparison

The two changes still reachable as parameters, measured end to end on this
machine. Note what this **cannot** show: the vectorised matrices, the flattened
bookkeeping and the fused objective are unconditional now, so they are present in
*both* arms. This comparison therefore isolates the solver and the tolerance
only; the full arc is §2.

```{r wholefit, eval=LIVE}
wf_args <- if (identical(params$scale, "spx")) {
  list(log_price = df_spx$logpris, fh = 500, hold_out = 500,
       lower = c(0.1, 6, -0.5, 0.8), upper = c(0.9, 13, -1e-14, 1e6),
       tc_init = n_spx + 250, m_init = 0.2, o_init = 12, num_searches = 1,
       mode = "MPL", mpl_cutoff = c(0.05, 0.1, 0.5))
} else {
  list(log_price = y_q, fh = 60, hold_out = 0,
       lower = c(0.1, 6, -0.03, 0.8), upper = c(0.9, 13, -1e-14, 1e6),
       tc_init = n_q + 30, m_init = 0.2, o_init = 12, num_searches = 1,
       mode = "MPL", mpl_cutoff = c(0.05, 0.1, 0.5))
}
run_wf <- function(method, factr) {
  a <- c(wf_args, list(beta_method = method, factr = factr))
  ts <- numeric(REPS); fit <- NULL
  for (i in seq_len(REPS)) {
    gc(FALSE)
    ts[i] <- system.time(fit <- suppressWarnings(do.call(fit_lppls, a)))[["elapsed"]]
  }
  list(t = stats::median(ts), tc = fit$fit[[1]]$tc)
}
w_old <- run_wf("symengine", 1e-08)
w_new <- run_wf("crossprod", 1e7)
wf <- data.frame(
  configuration = c("symengine solver, factr = 1e-08 (original)",
                    "crossprod solver, factr = 1e7 (current default)"),
  seconds = c(w_old$t, w_new$t),
  tc_hat = sprintf("%.8f", c(w_old$tc, w_new$tc)))
wf$speedup <- c("--", sprintf("%.1fx", w_old$t / w_new$t))
names(wf) <- c("configuration", "seconds", "estimated tc", "speed-up")
wf
```

```{r wholefit-note, results="asis", eval=LIVE}
cat(sprintf("Workload: `%s`. The two estimates agree to the precision the model reports.\n",
            params$scale))
```

# What was declined, and why

**Internal parallelism.** Distributing the `tc` loop over cores measured 4.76x at
8 cores with bit-identical results, and was still declined: the loop is not
side-effect free (`out_of_range_tracker`), random restarts break reproducibility
under `mclapply`, efficiency collapses past 4 cores on this hardware, and a
planned sweep layer over many series is a wider and safer parallel axis. Cores
spent inside `fit_lppls()` are cores taken from the sweep.

**A compiled `src/`.** Working Rcpp prototypes exist and were measured
(`misc/rcpp_survey.md`, `misc/rcpp_survey_prototypes.cpp`). The headline ratios
looked decisive — 13.9x for `compute_X_matrix`, 33.4x for `compute_H_matrix` —
but those were measured against the *interpreted loops*. Vectorised R recovers
most of that for free, and once it does, the whole remaining gain from compiling
is about **1.2x**. That did not justify a permanent `src/` directory, a C++
toolchain for source installs and more CRAN platforms to satisfy, so it was
deferred rather than adopted. The measurements and the reasoning are in
`misc/rcpp_survey.md`.

# Which changes altered the arithmetic

Some changes are bit-identical; others shift the trailing digits. None moves a
reported quantity, but the distinction matters for anyone comparing cached fits.

```{r invariance}
data.frame(
  change = c("crossprod solver (from symengine)",
             "factr = 1e7",
             "vectorised MPL matrices",
             "crossprod(X) for X'X",
             "flattened bookkeeping",
             "trace replay length",
             "fused objective"),
  bitwise = c("no", "no", "YES", "YES", "YES", "YES (fit unaffected)", "no"),
  effect = c("estimates differ ~1e-14",
             "estimates unchanged at reported precision; convergence code 52 -> 0",
             "matrices identical",
             "identical, and exactly symmetric",
             "whole fit object identical",
             "plot only; fewer, non-duplicated points",
             "SSE differs ~1 ulp; tc in the 12th digit"),
  stringsAsFactors = FALSE
)
```

A calibration is bit-reproducible only against the same `beta_method` and
`factr`. `beta_method = "symengine"` reproduces the thesis calibration.

# Method

Three things that cost real time to learn, recorded so they need not be
rediscovered:

1. **Measure in a fresh R process.** Calling `pkgload::unload()` and then
   `load_all()` on a second copy in one session inflated the MPL stage roughly
   fivefold — 19.1 s against 3.7 s for the same code. GC pressure was ruled out
   explicitly (nothing retained / all fits retained / 20,000 ballast objects:
   8.51, 8.60, 8.62 s).
2. **Run a control before attributing anything.** Every figure comparison in this
   work rendered twice with nothing changed first, to establish that the render
   was deterministic.
3. **`Rprof()` crashes R during an MPL fit on macOS.** It is R's sampling
   profiler colliding with Apple Accelerate's `DGEMM`, not a package fault; it
   reproduces in four lines of base R with no packages loaded. Use
   `options(matprod = "internal")` for the profiling session. Full account in
   `misc/rprof_mpl_crash.md`.

# Reproducing this report

```r
rmarkdown::render("misc/performance.Rmd",
                  params = list(live = TRUE, scale = "quick"))
```

`scale = "spx"` runs the whole-fit comparison on the real S&P 500 calibration
instead of the small synthetic series; expect it to take a few minutes, most of
it in the symengine arm. `live = FALSE` renders the recorded numbers alone.
