---
title: "The `lpplsF` package: A guided tour"
subtitle: "Fitting, diagnostics and Lagrange regularization for the LPPLS bubble model"
author: "lpplsF"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: kable
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    df_print: kable
params:
  recompute: false   # FALSE (default) = load cached fits from the folder (fast).
                     # TRUE            = re-run the single-model calibrations and refresh the cache.
---

```{r setup, include=FALSE}
## ── Reproducibility switch (set in the YAML header) ──────────────────────────
##   params$recompute = FALSE : load the cached single-model fits (fast; default).
##                    = TRUE  : re-run fit_e008 / fit_e009 / spx_fit and refresh the
##                              cache. The cheap SSE surfaces are always computed live;
##                              the multi-fit noise / rolling tables are always loaded
##                              from their pre-computed .rds (see "Reproducibility").
RECOMPUTE <- isTRUE(params$recompute)

## Pre-computed .rds live in this subdirectory (both saved and loaded here).
DATA_DIR <- "reproduce_thesis_plots_data"
dpath    <- function(f) file.path(DATA_DIR, f)
dir.create(DATA_DIR, showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.4, dpi = 110, fig.align = "center")

## Load the package from source, wherever this file is knit from.
pkg_root <- "."
for (up in c(".", "..", file.path("..", ".."), file.path("..", "..", ".."))) {
  if (file.exists(file.path(up, "DESCRIPTION"))) { pkg_root <- up; break }
}
suppressMessages(pkgload::load_all(pkg_root, quiet = TRUE))
suppressMessages(library(ggplot2))

## ── Format-aware SSE surface ─────────────────────────────────────────────────
## The package draws the SSE landscape as an interactive plotly 3-D surface
## (contour_plot_sse(..., sp = TRUE)$surface_plot, camera eye = (1.1, -1.3, 0.1)).
## plotly is an HTML widget, so for PDF output we redraw the same SSE grid as a
## static base-R persp() surface. `show_surface()` picks the right one for the
## current output format; both use the identical grid and axis labels.
grid_sse <- function(log_p, t, par, vars, lower, upper, mode = 0, n = 101) {
  x <- seq(lower[1], upper[1], length.out = n)
  y <- seq(lower[2], upper[2], length.out = n)
  f <- function(xx, yy) {
    xy <- stats::setNames(list(xx, yy), c(vars$x, vars$y))
    SSE(par = c(par, xy), log_p = log_p, t = t, mode = mode)
  }
  z <- t(outer(x, y, Vectorize(f)))               # dim [ny, nx]; z[j, i] = f(x[i], y[j])
  list(x = x, y = y, z = z, xlab = vars$x, ylab = vars$y)
}

persp_sse <- function(g, theta = -35, phi = 22) {
  Z   <- t(g$z)                                   # -> [nx, ny] for persp
  nrz <- nrow(Z); ncz <- ncol(Z)
  zf  <- (Z[-1, -1] + Z[-1, -ncz] + Z[-nrz, -1] + Z[-nrz, -ncz]) / 4  # facet height
  ## Match the plotly HTML surface (its default colorscale is viridisLite::viridis,
  ## the thesis colours); fall back to base viridis if viridisLite is unavailable.
  pal <- if (requireNamespace("viridisLite", quietly = TRUE)) {
    viridisLite::viridis(64)
  } else {
    grDevices::hcl.colors(64, "viridis")
  }
  graphics::persp(g$x, g$y, Z, theta = theta, phi = phi,
                  col = pal[cut(zf, 64, labels = FALSE)], border = NA, shade = 0.35,
                  xlab = g$xlab, ylab = g$ylab, zlab = "",
                  ticktype = "detailed", nticks = 4,
                  main = sprintf("SSE w.r.t. %s and %s", g$xlab, g$ylab))
}

show_surface <- function(log_p, t, par, vars, lower, upper, mode = 0, ylab = vars$y) {
  if (knitr::is_latex_output()) {
    g <- grid_sse(log_p, t, par, vars, lower, upper, mode); g$ylab <- ylab
    persp_sse(g)
  } else {
    p <- contour_plot_sse(log_p = log_p, t = t, par = par, vars = vars,
                          lower = lower, upper = upper,
                          cp = FALSE, sp = TRUE, mode = mode)$surface_plot
    if (!identical(ylab, vars$y)) {
      p <- plotly::layout(p, scene = list(yaxis = list(title = ylab)))
    }
    p
  }
}
```

`lpplsF` fits the **log-periodic power-law singularity (LPPLS)** model of Sornette
and co-authors — a model of the faster-than-exponential, log-periodically
decorated price growth that can precede a financial-bubble crash. This document
is a **guided tour of the package**: every calibration, diagnostic and plot it
offers, demonstrated on the exact examples from the master's thesis the package
was written for. It doubles as a **reproduction** of the thesis figures, so the
implementation can be checked against them.

**How to read it.** Each section builds on the previous one, from a single
calibration to rolling-window bubble-onset detection. Where the package output
differs from a thesis figure, a boxed note explains why — several are thesis
bugs the package corrects (flagged `T-5`…`T-8`, cross-referenced in
`misc/change_list.md`).

**Reproducibility.** The expensive single-model calibrations are cached next to
this file, so the document knits in seconds. Knit with

```r
rmarkdown::render("reproduce_thesis_plots.Rmd", params = list(recompute = TRUE))
```

to re-run them from scratch (a few minutes). The multi-fit noise and
rolling-window tables (tens of minutes to hours of CPU) are shipped precomputed
and loaded either way — a reader who wants to regenerate those can, using the
specification given in each section (see [Reproducibility and the cache]).

**Two output formats.** In HTML the SSE landscapes are interactive `plotly`
surfaces (drag to rotate; camera preset to the thesis viewpoint
`eye = (1.1, -1.3, 0.1)`); in PDF the same grids are drawn as static base-R
`persp()` surfaces.

# The model and its calibration landscape

The core generator is `eval_lppls()`. A single LPPLS trajectory is a
deterministic function of a critical time $t_c$, the power-law exponent $m$, the
log-periodic angular frequency $\omega$, and linear coefficients $A, B, C_1, C_2$:

```{r intro-eval}
str(eval_lppls(1:5, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001,
               tc = 1001, m = 0.6, omega = 8))
```

Calibration means minimizing the sum of squared errors (SSE) between the model
and observed log-prices over the nonlinear parameters $(t_c, m, \omega)$; the
linear parameters are profiled out. Before fitting, it helps to *see* that SSE
landscape. `contour_plot_sse()` evaluates the SSE over a grid of two parameters
(holding the rest fixed) and returns an interactive surface; here we wrap it in
`show_surface()` so the same grid also renders statically for PDF.

## Simulated LPPLS(0) data (Figure 6.1)

Thesis figure `fig:lpplsalt1`. A single LPPLS(0) process (the Filimonov–Sornette
parameterization, `mode = 0`) of length $N = 1000$ plus Gaussian noise
($\sigma = 0.05$); the SSE is evaluated on the first 910 observations (the
critical time $t_c = 1001$ lies beyond the fitting window).

| $t_c$ | $m$ | $\omega$ | $A$ | $B$ | $C_1$ | $C_2$ |
|----:|----:|------:|---:|------:|------:|-----:|
| 1001 | 0.6 | 8 | 8 | -0.025 | 0.0015 | 0.001 |

```{r fig61-simulate}
set.seed(3630)
sigma <- 0.05
N     <- 1000
tc    <- N + 1                      # 1001
sim   <- vapply(1:N, function(i) {
  eval_lppls(i, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001,
             tc = tc, m = 0.6, omega = 8) + rnorm(1) * sigma
}, numeric(1))

log_p <- sim[1:910]                 # fitting window
t_obs <- (1:N)[1:910]
```

### SSE over $m$ and $\omega$ (`lppls0_mo`)

Flat/low for small $m$, a weak depression near $m = 0.6$, then a steep rise for
larger $m$; only small oscillating variation along $\omega$.

```{r fig61-mo}
show_surface(
  log_p = log_p, t = t_obs,
  par   = list(tc = 1001, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001),
  vars  = list(x = "m", y = "omega"),
  lower = c(0.1, 6), upper = c(0.9, 13), mode = 0
)
```

### SSE over $t_c$ and $m$ (`lppls0_tc_m`)

Flat for small $m$, rising strongly for larger $m$; the variation along $t_c$ is
small for small $m$.

```{r fig61-tc-m}
show_surface(
  log_p = log_p, t = t_obs,
  par   = list(omega = 8, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001),
  vars  = list(x = "tc", y = "m"),
  lower = c(N - 90 + 1, 0.1), upper = c(N + 90, 0.9), mode = 0
)
```

### SSE over $t_c$ and $\omega$ (`lppls0_tc_o`)

Many strong oscillations along the $\omega$ direction — the source of the
multi-modal, hard-to-optimize likelihood that motivates the profile methods below.

```{r fig61-tc-o}
show_surface(
  log_p = log_p, t = t_obs,
  par   = list(m = 0.6, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001),
  vars  = list(x = "tc", y = "omega"),
  lower = c(N - 90 + 1, 1), upper = c(N + 90, 20), mode = 0
)
```

## Higher-order Landau variants (Figure 6.2)

Thesis figure `fig:lpplsalt2`. The package also implements the 1st- and 2nd-order
Landau expansions (`mode = 1`, `mode = 2`). In those modes $C_2$ plays the role
of the phase $\phi$, so a "$\phi$ axis" is produced by scanning $C_2$ (we relabel
the axis accordingly via `show_surface(..., ylab = "phi")`).

### $t_c$ and $\phi$, 1st-order Landau (`lppls1_tc_phi`)

This surface uses the **$N = 1000$ mode-0 data from Figure 6.1**, examined under
the mode-1 $(t_c, \phi)$ landscape — modes 0 and 1 are equivalent
parameterizations of the same 1st-order Landau LPPL. The phase periodicity
($2\pi$) produces three local minima across $\phi \in [1, 20]$, plus a boundary
solution at small $\phi$.

> **Thesis-input fix.** The original generator `surface_plots.Rmd` accidentally
> fed this plot the $N = 2000$ mode-2 data (an object had been reassigned by an
> earlier chunk), giving SSE $\approx 1000$ over $t_c \in [1911, 2090]$ — not the
> thesis figure. The package is correct; only the input needed fixing. Using the
> intended $N = 1000$ data reproduces the figure ($t_c \in [911, 1090]$).

```{r fig62-lppls1-tc-phi}
show_surface(
  log_p = log_p, t = t_obs,
  par   = list(m = 0.6, omega = 8, A = 8, B = -0.025, C1 = 0.0015),
  vars  = list(x = "tc", y = "C2"),
  lower = c(N - 90 + 1, 1), upper = c(N + 90, 20), mode = 1, ylab = "phi"
)
```

### A 2nd-order Landau simulation

The 2nd-order model adds a pre-critical regime of length $\Delta t = 500$ with
secondary frequency $\Delta\omega = 0$. SSE is evaluated on the first 1910 of
$N = 2000$ observations.

| $t_c$ | $m$ | $\omega$ | $A$ | $B$ | $C_1$ | $C_2$ | $\Delta t$ |
|----:|----:|------:|---:|------:|------:|-----:|----:|
| 2001 | 0.6 | 8 | 8 | -0.025 | 0.0015 | 0.001 | 500 |

```{r fig62-simulate}
set.seed(3630)
N2  <- 2000
tc2 <- N2 + 1
sim2 <- vapply(1:N2, function(i) {
  eval_lppls(i, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001,
             tc = tc2, m = 0.6, omega = 8, mode = 2, T1 = 500) + rnorm(1) * sigma
}, numeric(1))
log_p2 <- sim2[1:1910]
t_obs2 <- (1:N2)[1:1910]
```

### $t_c$ and $\phi$, 2nd-order Landau (`lppls2_tc_phi`)

Note the wider $t_c$ axis: the pre-critical regime lets parameters wander outside
the usual LPPLS filter values.

```{r fig62-lppls2-tc-phi}
show_surface(
  log_p = log_p2, t = t_obs2,
  par   = list(m = 0.6, omega = 8, A = 8, B = -0.025, C1 = 0.0015),
  vars  = list(x = "tc", y = "C2"),
  lower = c(N2 - 90 + 1, 1), upper = c(N2 + 200, 20), mode = 2, ylab = "phi"
)
```

### $t_c$ and $\omega$, 2nd-order Landau (`lppls2_tc_o`)

Even more rugged than the Filimonov variant: narrow, deep ravines and strong
oscillations along $\omega$.

```{r fig62-lppls2-tc-o}
show_surface(
  log_p = log_p2, t = t_obs2,
  par   = list(m = 0.6, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001),
  vars  = list(x = "tc", y = "omega"),
  lower = c(N2 - 90 + 1, 1), upper = c(N2 + 200, 20), mode = 2
)
```

# Fitting a single model

`fit_lppls()` is the workhorse. With `mode = "MPL"` it calibrates by **modified
profile likelihood**: for each candidate $t_c$ in the search window it profiles
out $(m, \omega)$ (and the linear parameters), then builds a likelihood interval
for $t_c$. We calibrate two series from thesis figure `fig:fit` (§8.1): one with
Gaussian noise (`M1`, experiment `e008`), one with AR(1)-GARCH(1,1) noise (`M2`,
`e009`). Each holds out the last 200 points, fits on $[1, 1800]$, and searches
$t_c$ over $[1801, 2100]$; the true $t_c = 2001$.

```{r figfit-helper}
## One-row tidy estimate table from a fit_lppls() result.
est_row <- function(fit) {
  e <- fit$fit[[1]]
  data.frame(tc = e$tc, m = e$m, omega = e$omega,
             A = unname(e$A), B = unname(e$B),
             C1 = unname(e$C1), C2 = unname(e$C2), D = unname(e$D))
}

## The calibration shared by M1 and M2 (matches thesis e008 / e009). fp/pp/tp
## ask fit_lppls to also attach the fit plot, the per-tc parameter plot, and the
## three BFGS trace plots to the returned object.
fit_call <- function(log_price) {
  fit_lppls(
    log_price = log_price, fh = 300, hold_out = 200,
    lower = c(0.1, 6, -0.03, 0.8), upper = c(0.9, 13, -1e-14, 1e6),
    tc_init = 1950, m_init = 0.2, o_init = 12, num_searches = 1,
    mode = "MPL", mpl_cutoff = c(0.05, 0.1, 0.5),
    fp = TRUE, pp = TRUE, tp = c(1, 1, 1), tp_id = 1
  )
}
```

The Gaussian series is regenerated from `set.seed(3630)`; the GARCH series is
loaded from `e009_garch_series.rds` (the exact thesis path — `rugarch`'s output
is version-sensitive and cannot be regenerated bit-for-bit).

```{r figfit-data}
set.seed(3630)
Nf <- 2000
gauss_series <- vapply(1:Nf, function(i) {
  eval_lppls(i, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001,
             tc = Nf + 1, m = 0.6, omega = 8) + rnorm(1) * sigma
}, numeric(1))
garch_series <- readRDS(dpath("e009_garch_series.rds"))$log_p
```

```{r figfit-compute, eval=RECOMPUTE}
## ~1 minute each (300 profile-likelihood optimizations per series).
fit_e008 <- fit_call(gauss_series)
fit_e009 <- fit_call(garch_series)
```

```{r figfit-save, eval=RECOMPUTE, echo=FALSE}
saveRDS(fit_e008, dpath("fit_e008.rds"))
saveRDS(fit_e009, dpath("fit_e009.rds"))
```

```{r figfit-load, eval=!RECOMPUTE, echo=FALSE}
fit_e008 <- readRDS(dpath("fit_e008.rds"))
fit_e009 <- readRDS(dpath("fit_e009.rds"))
```

## The fit plot

In each plot the blue line is the simulated series, the red curve the estimated
LPPLS process (up to the estimated $t_c$), the green dashed line marks
$T_2 = 1800$, and the red dashed line the estimated $t_c$. The package reproduces
the thesis estimates (Table `lpplsim1`) to machine precision.

> The package fit plot uses English axis labels and a native ggplot legend
> (`sim data` / `LPPLS fit` / `T2` / `tc_hat`); the thesis figures used Danish
> labels and a manually composited legend.

### M1 — Gaussian noise (`e008`)

```{r figfit-e008, fig.width=6, fig.height=4}
fit_e008$fit_plot
```

```{r figfit-e008-est}
knitr::kable(est_row(fit_e008), digits = 5,
             caption = "M1 (Gauss) estimate — cf. Table lpplsim1 (tc_hat = 1996).")
```

### M2 — AR(1)-GARCH(1,1) noise (`e009`)

```{r figfit-e009, fig.width=6, fig.height=4}
fit_e009$fit_plot
```

```{r figfit-e009-est}
knitr::kable(est_row(fit_e009), digits = 5,
             caption = "M2 (GARCH) estimate — cf. Table lpplsim1 (tc_hat = 1980).")
```

## Reading the fit: `print()` diagnostics

A calibration can look perfectly good as a plot and still be untrustworthy. The
`print()` method for `fit_lppls()` reports the estimate together with the checks
that decide whether it means anything — this is the fastest way to find out
whether a fit is worth interpreting.

```{r figfit-print}
fit_e008
```

Line by line:

* **Bounds.** $(m, \omega)$ are optimised inside a box. If the optimum sits *on*
  a wall, the SSE is still falling in that direction and the estimate is a
  constrained one — the regularity conditions behind the modified profile
  likelihood do not hold, and its intervals mean nothing. The check is reported
  twice: for the point estimate, and as a count across all the per-$t_c$
  conditional fits. The second matters on its own, because an interior point
  estimate can still rest on a profile that is pinned nearly everywhere.
* **Basins.** Each candidate $t_c$ gets its own $(m, \omega)$ search, and with
  few restarts some land in a worse local optimum than others. Distance from the
  best SSE will not detect this — the conditional SSE *should* rise away from the
  optimal $t_c$, so a healthy profile also has fits well above the minimum. What
  betrays a failed search is a **gap**: the sorted SSEs falling into a good
  cluster and a distinctly worse one. The reported number is the largest gap
  between consecutive sorted values as a share of the total spread, which is
  scale-free and near zero for a smooth profile.
* **Filters.** How many candidate fits were rejected for $B \ge 0$ or for
  damping below its floor — the two constraints `fit_lppls()` enforces
  internally, otherwise visible only as warnings while the fit runs.
* **MPL.** How much of the $t_c$ grid the modified profile likelihood is defined
  on. `compute_mpl()` returns `NA` wherever $\det(X'X - H) \le 0$, which at a
  constrained optimum carries no guarantee of sign.

A `Diagnostics:` block of `!` lines appears only when something is wrong, so a
clean fit prints none — as here. §5 shows one that prints three.

```{r figfit-print-e009}
fit_e009
```

## Per-$t_c$ parameter estimates (`param`)

Passing `pp = TRUE` to `fit_lppls()` attaches a parameter plot: each estimated
parameter (and the objective) as a function of the candidate $t_c$, with the
objective minimum ($\hat t_c$) marked. $C_1$ and $C_2$ trace near-sinusoidal
curves that distort far from the true $t_c$ — more so under GARCH noise. No extra
calibration runs; these reuse the fits above.

### M1 — Gaussian noise

```{r param-e008, fig.width=5, fig.height=8}
fit_e008$param_plot
```

### M2 — AR(1)-GARCH(1,1) noise

```{r param-e009, fig.width=5, fig.height=8}
fit_e009$param_plot
```

# Reading the optimizer: BFGS traces

Thesis figures `trace` / `fit` / `tracebo` (§8.2). With `tp = c(1, 1, 1)`,
`fit_lppls()` attaches three trace plots: each shows the SSE landscape for a pair
of parameters at the estimated $t_c$, overlaid with the BFGS path — red dot at
the start $(m_{\text{init}}, \omega_{\text{init}}) = (0.2, 12)$, green diamond at
the estimate.

## $(m, \omega)$ topology

The governing topology for calibration: a single interior minimum BFGS finds
easily. The GARCH minimum sits slightly further from the true $(0.6, 8)$.

```{r trace-mo-e008, fig.width=6, fig.height=4}
fit_e008$trace_plot_mo   # Gauss
```

```{r trace-mo-e009, fig.width=6, fig.height=4}
fit_e009$trace_plot_mo   # GARCH
```

## $(B, m)$ topology

Very steep for large $m$ and large negative $B$ (note the SSE scale $\sim 10^8$),
flat elsewhere. $B$ is in fact determined by $(m, \omega)$; this view treats it
as free, for illustration.

```{r trace-bm-e008, fig.width=6, fig.height=4}
fit_e008$trace_plot_bm   # Gauss
```

```{r trace-bm-e009, fig.width=6, fig.height=4}
fit_e009$trace_plot_bm   # GARCH
```

## $(B, \omega)$ topology

SSE varies only in small waves along $\omega$.

> **Thesis bug (T-5).** The thesis `tracebo` figures are inconsistent: their
> start $B$ ($\approx -0.4$) does not match the $(B, m)$ figures' start $B$
> ($\approx -1.04$), though both are the same `init_B`, and the drawn path does
> not converge to the estimate. The package computes the optimization path once
> and shares it across all three trace plots, so the $(B, \omega)$ trace here is
> self-consistent (same `init_B` as $(B, m)$; path ends at the estimate).

```{r trace-bo-e008, fig.width=6, fig.height=4}
fit_e008$trace_plot_bo   # Gauss (corrected)
```

```{r trace-bo-e009, fig.width=6, fig.height=4}
fit_e009$trace_plot_bo   # GARCH (corrected)
```

# Robustness to noise

Thesis figures `fig:noise_param` / `fig:gauss` / `fig:garch` (§8.3). Four
experiments calibrate the model at **20 noise levels** each: `e010` (M1, Gauss),
`e012` (M2, Gauss), `e011` (M1, GARCH), `e011b` (M2, GARCH), where M2 is M1 with
$C_1, C_2$ negated (opposite phase). Each level is an MPL calibration
(`hold_out = 100`, `fh = 150`).

That is **80 calibrations ($\approx$ 45 min)**, so the per-level result tables
are shipped precomputed (`noise_*.rds`) and loaded here; see
[Reproducibility and the cache].

> **Provenance.** All four sweeps were produced with the current package. The two
> Gaussian series regenerate from `set.seed(3630)`; the two GARCH series do *not*
> (they came from `rugarch::ugarchpath`, whose simulated path is
> version-dependent), but the 20 original paths survive alongside the thesis
> experiments, so no re-simulation is needed: `e011` refits the saved path
> directly, and `e011b` — built from the same seed and spec — refits
> `LPPLS(-C_1, -C_2)` plus the noise recovered as
> `saved path - LPPLS(+C_1, +C_2)`. Both reproduce the previously cached point
> estimates exactly, which confirms the recovered series are the originals.

> **Thesis bug (T-6).** The thesis $t_c$ likelihood intervals are too narrow: it
> computed the per-$t_c$ profile variance $s_{t_c}$ with $m, \omega$ held at the
> *global* optimum instead of profiling them per $t_c$. The package fixes this
> (verified to the digit), so the interval bands below are **wider (correct)**.
> The point estimates and the parameter-vs-$\sigma$ plots are unaffected.

```{r noise-setup}
noise_e010  <- readRDS(dpath("noise_e010.rds"))   # M1 Gauss
noise_e012  <- readRDS(dpath("noise_e012.rds"))   # M2 Gauss
noise_e011  <- readRDS(dpath("noise_e011.rds"))   # M1 GARCH
noise_e011b <- readRDS(dpath("noise_e011b.rds"))  # M2 GARCH

## Parameter estimates vs noise level.
param_vs_sd <- function(tab) {
  d <- tidyr::gather(tab[, c("m", "omega", "A", "B", "C1", "C2", "sd")],
                     key = "param", value = "estimate", -sd)
  ggplot(d, aes(sd, estimate)) +
    geom_line(linewidth = 0.3) +
    facet_wrap(~param, scales = "free_y", ncol = 1) # +
    # theme_minimal()
}

## tc point estimate (PLE/MPLE) + nested likelihood-interval bands vs noise level.
li_vs_sd <- function(tab) {
  cols <- c(LI_05 = "green", LI_10 = "orange", LI_50 = "red",
            MPLE = "cyan", PLE = "black")
  ggplot(tab, aes(sd, tc)) +
    geom_ribbon(aes(ymin = LI5l,  ymax = LI5u,  color = "LI_05"),
                fill = "grey50", alpha = 0.2, key_glyph = "path") +
    geom_ribbon(aes(ymin = LI10l, ymax = LI10u, color = "LI_10"),
                fill = "grey50", alpha = 0.2, key_glyph = "path") +
    geom_ribbon(aes(ymin = LI50l, ymax = LI50u, color = "LI_50"),
                fill = "grey50", alpha = 0.2, key_glyph = "path") +
    geom_line(aes(color = "PLE"), linewidth = 1) +
    geom_line(aes(y = tc_hat_mpl, color = "MPLE"), linewidth = 0.25) +
    scale_color_manual(name = NULL, values = cols, breaks = names(cols)) +
    labs(x = "noise sd", y = quote(hat(t)[c])) +
    theme_minimal()
}
```

## Parameter estimates vs noise level

$\omega, C_1, C_2$ (the log-periodic parameters) are the most noise-sensitive.

### Gauss noise (`e010`)

```{r noise-sd-gauss, fig.width=5, fig.height=8}
param_vs_sd(noise_e010)
```

### AR(1)-GARCH(1,1) noise (`e011`)

```{r noise-sd-garch, fig.width=5, fig.height=8}
param_vs_sd(noise_e011)
```

## $t_c$ likelihood intervals vs noise level

True $t_c = 2001$. Bands are the corrected (wider) intervals.

### Gauss, M1 (`e010`) and M2 opposite phase (`e012`)

```{r noise-li-e010, fig.width=6, fig.height=4}
li_vs_sd(noise_e010)
```

```{r noise-li-e012, fig.width=6, fig.height=4}
li_vs_sd(noise_e012)
```

### GARCH, M1 (`e011`) and M2 opposite phase (`e011b`)

```{r noise-li-e011, fig.width=6, fig.height=4}
li_vs_sd(noise_e011)
```

```{r noise-li-e011b, fig.width=6, fig.height=4}
li_vs_sd(noise_e011b)
```

# Rolling windows and Lagrange regularization

Thesis figure `fig:lagrange2` (§8.4). A bubble has a *beginning*; Lagrange
regularization (Demos & Sornette, 2019) tries to find it by rolling the
calibration start $T_1$ forward (fixed end $T_2$) and looking for the window
where the normalized residual $\chi^2_{np}(T_1)$ stops falling. The package
splits this into a reusable primitive and cheap post-processing:

* `lppls_rolling()` runs one calibration per $T_1$ (the expensive part);
* `lppls_lagrange()`, `rolling_tc_plot()`, `rolling_param_plot()`,
  `chi_sq_np_plot()`, `chi_sq_lambda_plot()` are all post-processing on it.

This only has bite when the series *has* an onset to find: in a pure LPPL process
every window is already LPPL, so $\chi^2_{np}$ is flat and regularization has
nothing to correct. So the example is a **dual-regime** series — a flat "ranging"
segment for $t < 600$ that switches to an LPPL process at $t = 600$
(`tc = 2001, m = 0.6, omega = 8, A = 8, B = -0.025, C1 = 0.0015, C2 = 0.001`;
Gaussian noise $\sigma = 0.05$, seed 3630). The rolling approaches the onset from
below: as $T_1$ climbs toward 600 each window sheds more of the pre-LPPL data and
the fit tightens, so $\chi^2_{np}$ decreases; once $T_1$ passes 600 the window is
pure LPPL again and the trend flattens. The rolling ($T_1 = 1 \to 1001$ step 10,
$T_2 = 1900$; 101 windows, $\approx$ 13 min) is shipped precomputed as
`lagrange_rolling_regime_gauss.rds`.

```{r regime-gauss-build}
## The dual-regime series: flat at the LPPL value at t = 600 for t < 600, then
## the LPPL process itself, with N(0, 0.05^2) noise throughout. The two segments
## draw their noise from separate seeds -- `seed` for the flat prefix, `seed + 1`
## for the bubble. Both set.seed() calls are load-bearing: drop either and the
## cached calibration no longer matches the series.
N_rg <- 2000
build_rg <- function(seed, onset = 600) {
  bub <- eval_lppls(onset:N_rg, 8, -0.025, 0.0015, 0.001, N_rg + 1, 0.6, 8)
  set.seed(seed + 1)
  c(rep(bub[1], onset - 1), bub) +
    c({ set.seed(seed); rnorm(onset - 1) * 0.05 }, rnorm(N_rg - onset + 1) * 0.05)
}

## Regenerated on both paths: it is instant, and the residual guard below needs
## the series even when the calibration comes from the cache.
lp_rg <- build_rg(3630)
```

```{r regime-gauss-compute, eval=RECOMPUTE}
## ~14 minutes: 101 windows, each a 150-candidate tc scan (MPL: the likelihood
## intervals rolling_tc_plot() needs).
roll_rg <- lppls_rolling(
  lp_rg,
  t1_seq = seq(1, 1001, by = 10),
  hold_out = 100,
  fh = 150,
  fit_args = list(
    lower = c(0.1, 6, -0.03, 0.8),
    upper = c(0.9, 13, -1e-14, 1e6),
    m_init = 0.2,
    o_init = 12,
    num_searches = 1,
    mode = "MPL",
    mpl_cutoff = c(0.05, 0.1, 0.5),
    factr = 1e7)
  )
```

```{r regime-gauss-save, eval=RECOMPUTE, echo=FALSE}
saveRDS(roll_rg, dpath("lagrange_rolling_regime_gauss.rds"))
```

```{r regime-gauss-load, eval=!RECOMPUTE, echo=FALSE}
roll_rg <- readRDS(dpath("lagrange_rolling_regime_gauss.rds"))
```

```{r regime-gauss-lagrange}
lag_rg <- lppls_lagrange(roll_rg, p = 0)
lag_rg
```

Optimal $T_1^* =$ **`r lag_rg$t1_star`** (true onset 600); $\lambda =$
`r signif(-lag_rg$slope, 3)` $> 0$ — a genuine bias trend (a pure LPPL series
would give $\lambda \approx 0$).

```{r regime-gauss-window}
k_rg  <- which(roll_rg$t1_seq == lag_rg$t1_star)
T1_rg <- roll_rg$t1_seq[k_rg]        # the calibration start (optimal T1*)
nc_rg <- roll_rg$table$n_cal[k_rg]
T2_rg <- T1_rg + nc_rg - 1           # 1900, the calibration end
f_rg  <- roll_rg$fits[[k_rg]]
```

## The generated series and its fit

Blue the generated series, red the fitted LPPLS process up to $\hat t_c$, dashed
markers at the calibration bounds and $\hat t_c$. The series spans two regimes and
the fit is calibrated on only the later part, so the red curve starts at $T_1^*$ —
the flat regime is shown but never fitted. `lppls_rolling()` stores the fit in
window-local time ($t = 1$ at $T_1$), so $\hat t_c$ is shifted by the window offset
to draw it on the global clock.

```{r regime-gauss-fit, fig.width=6, fig.height=4}
f_glob    <- f_rg
f_glob$tc <- f_rg$tc + (T1_rg - 1)   # window-local -> global clock

if ("T1" %in% names(formals(fit_plot))) {
  ## fit_plot() can bound the fitted curve at T1 (package patch Part 1).
  fit_plot(f_glob, seq_along(lp_rg), lp_rg, T1 = T1_rg, T2 = T2_rg)
} else {
  ## Current dev: fit_plot() starts the curve at the first plotted point, so plot
  ## the calibrated window, then add the earlier regime and a T1* marker on top.
  fit_plot(f_glob, T1_rg:roll_rg$n_full, lp_rg[T1_rg:roll_rg$n_full], T2 = T2_rg) +
    geom_line(data = data.frame(ID = seq_len(T1_rg - 1),
                                log_p = lp_rg[seq_len(T1_rg - 1)]),
              aes(ID, log_p), color = "royalblue1") +
    geom_vline(xintercept = T1_rg, color = "orange", linetype = "dashed")
}
```

## Rolling-$T_1$ estimates (`fig:rullendeT1`)

How the $t_c$ estimate and the six parameters move as $T_1$ rolls across the
onset. For $T_1$ well below 600 the window still contains ranging data and the
estimates wander; as $T_1$ approaches and passes 600 the window becomes
predominantly, then purely, LPPL and $\hat t_c$ settles near the true 2001.

```{r regime-gauss-tc, fig.width=6, fig.height=4}
rolling_tc_plot(roll_rg)
```

```{r regime-gauss-params, fig.width=5, fig.height=8}
rolling_param_plot(roll_rg)
```

## Lagrange regularization (`fig:lagrange2`)

$\chi^2_{np}(T_1)$ with the OLS bias trend (red), then the regularized objective
$\chi^2_{\lambda^*}(T_1)$ with the optimal $T_1^*$ (green):

```{r regime-gauss-chi-np, fig.width=6, fig.height=4}
chi_sq_np_plot(lag_rg)
```

```{r regime-gauss-chi-lambda, fig.width=6, fig.height=4}
chi_sq_lambda_plot(lag_rg)
```

## Residuals at $T_1^*$

```{r regime-gauss-guard, echo=FALSE}
## Load-path integrity: the regenerated series must reproduce the cached
## calibration exactly (residual SSE == stored sse_cal).
res_rg <- lp_rg[T1_rg:roll_rg$n_full][seq_len(nc_rg)] -
  eval_lppls(seq_len(nc_rg), f_rg$A, f_rg$B, f_rg$C1, f_rg$C2,
             f_rg$tc, f_rg$m, f_rg$omega, mode = 0)
stopifnot(isTRUE(all.equal(sum(res_rg^2), roll_rg$table$sse_cal[k_rg])))
```

The residual QQ plot for the window at $T_1^*$. The residuals are near-normal:
the window at $T_1^*$ lies almost entirely inside the LPPL segment, so the shape
at a single window does not reveal the regime — the Lagrange signal is in the
$\chi^2_{np}$ trend *across* windows.

```{r regime-gauss-res, fig.width=6, fig.height=4}
residual_qq_plot(res_rg)
```

> **Is that decrease real, or a noise artifact?** That question — and the
> autocorrelation-robust `lr_trend_test()` that answers it — is the subject of a
> companion demo, `lr_demo.Rmd`, with the full investigation in
> `lagrange_inference_report.Rmd` (currently in the package's `misc/`
> directory; this location may change).

# Empirical example: the S&P 500

Chapter 9. The package ships the S&P 500 series (`inst/extdata/SPX.csv`); we
calibrate on rows 11800:14727 (2009-05-18 to 2020-12-31), fitting on $[1, 2428]$
(to 2019-01-08) with `hold_out = 500`, `fh = 500`.

```{r spx-data}
spx <- read.csv(system.file("extdata", "SPX.csv", package = "lpplsF"))
names(spx)[1] <- "Date"
spx$Date    <- as.Date(as.character(spx$Date), "%Y%m%d")
spx$logpris <- log(spx$spindx)
df_spx <- spx[11800:14727, ]; df_spx$ID <- seq_len(nrow(df_spx))
n_cal_spx <- nrow(df_spx) - 500    # 2428
```

```{r spx-compute, eval=RECOMPUTE}
## ~1-3 minutes (fh = 500 profile-likelihood optimizations).
spx_fit <- fit_lppls(
  log_price = df_spx$logpris, fh = 500, hold_out = 500,
  lower = c(0.1, 6, -0.5, 0.8), upper = c(0.9, 13, -1e-14, 1e6),
  tc_init = n_cal_spx + 250, m_init = 0.2, o_init = 12, num_searches = 1,
  mode = "MPL", mpl_cutoff = c(0.05, 0.1, 0.5),
  fp = TRUE, mpl_plot = TRUE, tp = c(1, 0, 0), pp = TRUE)
```

```{r spx-save, eval=RECOMPUTE, echo=FALSE}
saveRDS(spx_fit, dpath("spx_fit.rds"))
```

```{r spx-load, eval=!RECOMPUTE, echo=FALSE}
spx_fit <- readRDS(dpath("spx_fit.rds"))
```

```{r spx-est}
e <- spx_fit$fit[[1]]
```

Point estimate reproduces the thesis exactly: PLE $t_c =$
**`r as.character(df_spx$Date[round(e$tc)])`**, with $m =$ `r round(e$m, 2)`
(pinned at the upper boundary) and $\omega =$ `r round(e$omega, 2)`.

### What `print()` says about this one

§2 introduced the diagnostics on a healthy calibration, which printed no
warnings. This is the contrasting case, and it is worth reading before any of the
figures below:

```{r spx-print}
spx_fit
```

Three flags, and every caveat in the rest of this section follows from them. The
estimate is a **boundary solution**, and not marginally — $m$ is at its upper
bound in the point estimate *and* in every conditional fit, with
$\partial\,\mathrm{SSE}/\partial m < 0$ there, so the constraint is genuinely
active and the regularity conditions behind the modified profile likelihood do
not hold. Most of the conditional fits sit in a **distinctly inferior basin**,
which is why the MPL curve below is two-tiered rather than smooth, and why
`num_searches = 1` is the wrong setting for this series. And the MPL is
**undefined on part of the grid**, so the curve has gaps the likelihood intervals
cannot see across.

None of this is visible in the point estimate, which looks entirely ordinary.

## Fit plot (`fig:spx1`)

Here we build the fit plot directly (custom `Date` axis) rather than using the
attached `$fit_plot`, to show working with real dates.

```{r spx-fitplot, fig.width=8, fig.height=4}
lppls_fit  <- eval_lppls(1:n_cal_spx, e$A, e$B, e$C1, e$C2, e$tc, e$m, e$omega)
pred_ids   <- (n_cal_spx + 1):floor(e$tc)
lppls_pred <- eval_lppls(pred_ids, e$A, e$B, e$C1, e$C2, e$tc, e$m, e$omega)
dfp <- rbind(
  data.frame(Date = df_spx$Date,              logpris = df_spx$logpris, series = "data"),
  data.frame(Date = df_spx$Date[1:n_cal_spx], logpris = lppls_fit,      series = "fit"),
  data.frame(Date = df_spx$Date[pred_ids],    logpris = lppls_pred,     series = "prediction"))
dfp$series <- factor(dfp$series, levels = c("data", "fit", "prediction"))

ggplot(dfp, aes(Date, logpris, color = series)) +
  geom_line(linewidth = 0.3) +
  geom_vline(xintercept = df_spx$Date[n_cal_spx],   color = "royalblue", linetype = "dashed") +
  geom_vline(xintercept = df_spx$Date[round(e$tc)], color = "red",       linetype = "dashed") +
  scale_color_manual(name = NULL,
    values = c(data = "darkgray", fit = "royalblue", prediction = "red")) +
  labs(x = "Year", y = "log price") +
  theme_minimal() +
  theme(legend.position = "inside", legend.position.inside = c(0.01, 0.99),
        legend.justification.inside = c(0, 1),
        legend.background = element_rect(
          fill = grDevices::adjustcolor("white", alpha.f = 0.6), color = NA))
```

## Modified profile likelihood (`fig:spx2`)

The MPLE coincides with the PLE, at 2020-03-03. `compute_mpl()` returns `NA`
wherever $\det(X'X - H) \le 0$ — which at a constrained optimum carries no
guarantee of sign — hence the gaps the `MPL:` line counted above.

> **Thesis bug (T-8).** `filimonov_functions.R` builds the matrix $H$ of
> residual-weighted second derivatives without mirroring its $(\omega, C_1)$ and
> $(\omega, C_2)$ entries into the lower triangle. $H$ is symmetric by
> construction, so the thesis's $\det(X'X - H)$ is not the observed-information
> determinant; on this series its sign is wrong at 473 of the 500 grid points,
> which truncates the curve at $t_c = 2499$ and collapses all three
> likelihood-interval upper limits onto that point. The thesis appendix and
> Filimonov (2017) eq. (37) both give $H$ correctly, so this is an implementation
> slip, not a derivation error. The package computes $H$ symmetrically.

> **Diagnostic caveat.** The point estimate agrees with the PLE, but the bands are
> not inference — the `Bounds:` and `Basins:` lines above already say why. Two
> consequences the printed summary does not spell out. First, the basin split is
> what makes the curve **two-tiered** rather than smooth: the two tiers are
> $\approx 39$ log-likelihood units apart, so the plot below is not one curve but
> two.
>
> Second, and the reason to distrust the bands specifically: the likelihood
> intervals here trace **where the optimiser succeeded**, not the curvature of the
> likelihood. Every grid point inside LI_05 is a good-basin fit and every point
> outside it is not; its lower edge is exactly where the good-basin run containing
> the PLE begins, and its upper edge is the end of the *next* such run, which
> `range()` bridges across the gap between them. LI_05 and LI_10 coincide for the
> same reason. The relative-scale plot below makes this visible: only LI_50 has a
> crossing. **Read the MPLE, not the bands.**

```{r spx-mplplot, fig.width=5.5, fig.height=4}
spx_fit$mpl_plot
```

### The same curve on a relative scale

The absolute level of the log MPL carries no information. It is a log *density*
— here the residual $\sigma \approx 0.047$, so the density at its mode is
$\approx 8.6$ and the per-observation log is positive; multiplied by $n = 2428$
that is the $\approx 7000$ on the axis above. The formula also drops the additive
constant $-\tfrac{n}{2}\log 2\pi - \tfrac{n}{2}$, which cancels everywhere it is
used. What the likelihood intervals actually test is the *difference*
$R = \log\mathrm{MPL} - \max\log\mathrm{MPL}$ against $\log(\text{cutoff})$, and
`compute_mpl()` already returns it as `mpl_output$R`.

Plotting $R$ and restricting the view to the band the cutoffs live in turns the
intervals from something asserted into something visible. The package's
`create_mpl_plot()` draws the absolute curve, so this view is built here — a
fixed zoom would be wrong for the well-behaved fits of §2, whose curves lie
within a couple of log units of their maximum and need no zoom at all.

```{r spx-mpl-rel-helper}
## R = log MPL - max(log MPL), zoomed to the cutoff band. The vertical markers
## are the same five the package draws; the horizontal lines are the cutoffs
## themselves, so each interval can be read off where the curve crosses one.
mpl_rel_plot <- function(fit_obj, cutoff, pad = 1.2) {
  mo <- fit_obj$mpl_output
  ## fit_args is absent from caches written before fit_lppls() recorded it, so
  ## fall back to the tc grid, which starts at n + 1.
  n <- if (is.null(fit_obj$fit_args$n)) {
    min(fit_obj$fit[[2]]$tc) - 1
  } else {
    fit_obj$fit_args$n
  }

  cols <- c(PLE = "blue", MPLE = "blue", LI_05 = "green3",
            LI_10 = "orange", LI_50 = "red")
  ltys <- c(PLE = "solid", MPLE = "dashed", LI_05 = "dashed",
            LI_10 = "dotted", LI_50 = "dotdash")
  li_names <- c("LI_05", "LI_10", "LI_50")

  vmarks <- data.frame(xintercept = c(fit_obj$fit[[1]]$tc, mo$tc_hat_mpl),
                       series = c("PLE", "MPLE"))
  for (i in seq_along(mo$LI)) {
    if (!any(is.na(mo$LI[[i]])))
      vmarks <- rbind(vmarks,
                      data.frame(xintercept = mo$LI[[i]], series = li_names[i]))
  }
  present <- names(cols)[names(cols) %in% vmarks$series]
  vmarks$series <- factor(vmarks$series, levels = present)
  hmarks <- data.frame(yintercept = log(cutoff),
                       series = factor(li_names[seq_along(cutoff)],
                                       levels = present))

  ggplot(data.frame(tc = (n + 1):(n + length(mo$R)), R = mo$R),
         aes(x = tc, y = R)) +
    geom_point(size = 0.5, na.rm = TRUE) +
    geom_hline(data = hmarks,
               aes(yintercept = yintercept, color = series, linetype = series),
               linewidth = 0.4) +
    geom_vline(data = vmarks,
               aes(xintercept = xintercept, color = series, linetype = series),
               key_glyph = "path") +
    scale_color_manual(name = NULL, values = cols, breaks = present) +
    scale_linetype_manual(name = NULL, values = ltys, breaks = present) +
    coord_cartesian(ylim = c(min(log(cutoff)) - pad, 0.4)) +
    labs(x = "Critical Time (tc)", y = "Relative log MPL",
         title = "Modified Profile Likelihood (relative)") +
    theme_minimal() +
    lpplsF:::overlay_legend_theme(inside = c(0.99, 0.99),
                                  justification = c(1, 1))
}
```

```{r spx-mplplot-rel, fig.width=5.5, fig.height=4}
mpl_rel_plot(spx_fit, c(0.05, 0.1, 0.5))
```

Only `LI_50` has a crossing: the curve descends from the peak to about $-1.15$
and then stops, so it never reaches $\log 0.10$ or $\log 0.05$. Those two
intervals end where the good basin ends, which is the previous note made visible.

## The rest of the diagnostics

`print()` is the quick triage. Three further tools answer the questions it
raises: *where* the basins are, whether the estimate sits at a genuine minimum,
and whether the intervals are closed by the likelihood or by the grid.

### The basin split, drawn

`plot(fit, "basin")` shows every conditional fit's SSE against its $t_c$, split
at the gap `print()` reported. It needs nothing that was not already computed, so
it works even on a fit that requested no plots.

```{r spx-basin, fig.width=6, fig.height=3.6}
plot(spx_fit, "basin")
```

The same view of the well-behaved calibration from §2, for contrast — one cloud,
rising smoothly away from the optimal $t_c$, which is what a profile should look
like:

```{r e008-basin, fig.width=6, fig.height=3.6}
plot(fit_e008, "basin")
```

### What the basins are made of

The view above shows *that* the fits split, but not what they split on.
`"param_basin"` applies the same division to every estimated parameter rather
than only the objective, which is where the mechanism becomes visible.

```{r spx-param-basin, fig.width=6, fig.height=9.5}
plot(spx_fit, "param_basin")
```

```{r spx-basin-ranges}
## The split is the one summary() reports; here it is per parameter.
cut_at <- summary(spx_fit)$split_at[1]
pt     <- spx_fit$fit[[2]]
pt$basin <- ifelse(pt$value <= cut_at, "good", "inferior")
rng <- function(z) sprintf("%.3f - %.3f", min(z), max(z))
knitr::kable(
  aggregate(cbind(m, omega, D) ~ basin, data = pt, FUN = rng),
  caption = "Range of each nonlinear estimate within each basin.")
```

The two basins are separated almost entirely by $\omega$: the good one sits near
the bottom of the $[6, 13]$ search box, the inferior one near the top. $m$ is at
$0.9$ in **both**, so it plays no part in the split at all.

That separates two failures which the SSE view alone runs together. The pinned
$m$ is a **boundary** problem, and it affects every fit on the grid. The two
tiers are an **$\omega$ multimodality** problem, and it affects only which
optimum a single L-BFGS-B start happens to reach — a `num_searches` setting, not
a property of the data. The `C1`/`C2` facets tell the same story: the inferior
fits oscillate through a high-frequency regime while the good ones trace a smooth
low-frequency arc.

### Curvature at the estimate

The modified profile likelihood assumes the estimate is an interior minimum: its
correction uses $\det(X'X - H)$, the observed information, which is only
guaranteed positive at an interior stationary point. `lppls_curvature()` tests
that directly, and factorises the determinant to show where its sign comes from.

By the Schur complement, $\det(X'X - H)$ is the determinant of the
linear-parameter block — a crossproduct, so always positive — times the
determinant of the $2 \times 2$ Hessian of the $(m, \omega)$ profile. The sign is
therefore decided entirely by curvature in the two nonlinear parameters.

```{r spx-curvature}
lppls_curvature(spx_fit)
```

Read the classification and the bound note as two separate facts. The curvature
*is* positive definite here, so the $t_c$ the MPL reports is not an artefact of a
degenerate Hessian. But $m$ rests on its bound, so the gradient need not vanish:
the estimate is a constrained one whatever the curvature says. It is the
combination — sound curvature, active constraint — that makes the point estimate
usable and the intervals not.

### Everything at once: `summary()`

`summary()` returns the diagnostics as values rather than text, so they can be
tested against instead of parsed. Printing it gives the `print()` report plus the
curvature line and an interval breakdown.

```{r spx-summary}
s <- summary(spx_fit)
s$basin_gap
s$mpl$li_check[[1]]
```

That last list is the interval integrity check. `compute_mpl()` forms an interval
as `range()` of the $t_c$ whose $R$ exceeds $\log(\text{cutoff})$ — and `range()`
does not care whether that set has holes. When it does, the reported interval
contains grid points that are not themselves in it:

```{r spx-summary-print}
s
```

`LI(0.05)` and `LI(0.10)` each span more grid points than are above their cutoff:
the interval bridges the gap between two runs of good-basin fits. `LI(0.50)` is
contiguous. This is the same conclusion the relative-scale plot above shows
visually, arrived at from the numbers.

## Fit diagnostics (`fig:spxparam`)

The BFGS path slides to the $(m, \omega)$ boundary corner; the per-$t_c$
parameter panel shows $m$ pinned at 0.9 with jumps where it hits the boundary.

```{r spx-trace, fig.width=6, fig.height=4}
spx_fit$trace_plot_mo
```

```{r spx-param, fig.width=5, fig.height=8}
spx_fit$param_plot
```

## Residual QQ diagnostics (`fig:spxqq`)

Residuals of the SPX calibration ($t < \hat t_c$) against a normal and a fitted
Student-t reference, both via `residual_qq_plot()`.

```{r spx-qq-setup, echo=FALSE}
lppls_all <- eval_lppls(df_spx$ID, e$A, e$B, e$C1, e$C2, e$tc, e$m, e$omega)
res_spx   <- df_spx$logpris[!is.na(lppls_all)] - lppls_all[!is.na(lppls_all)]
```

```{r spx-qq-norm, fig.width=8, fig.height=4}
residual_qq_plot(res_spx)
```

```{r spx-qq-t, fig.width=8, fig.height=4}
residual_qq_plot(res_spx, "t")
```

## Lagrange regularization on the S&P 500 (`fig:spxlagrange1`)

Rolling $T_1 = 1 \to 501$ step 25 (21 windows), $T_2 = 2428$ fixed; shipped
precomputed as `lagrange_rolling_e016b.rds` (21-window MPL fit, $\approx$ 44 min).

The refit spec is the `spx-fit` call above with `t1_seq = seq(1, 501, 25)`,
`T2 = 2428`; `lppls_rolling()` sets each window's `tc_init` to
`n_cal + floor(fh / 2)`, so window 1 reproduces `spx_fit` exactly.

```{r e016b-setup}
roll_e016b <- readRDS(dpath("lagrange_rolling_e016b.rds"))
lag_e016b  <- lppls_lagrange(roll_e016b, p = 0)
lag_e016b
```

Optimal $T_1^* =$ **`r lag_e016b$t1_star`**
(= `r as.character(df_spx$Date[lag_e016b$t1_star])`), matching the thesis
(2009-09-02). Here $\beta_1 < 0$, $\lambda > 0$, so the Demos bias trend is
genuinely present. (The thesis $\chi^2$
*figures* still need regenerating: their $y$-values $\approx 182$ are the buggy
4th-moment artifact; the correct scale is $\approx 0.0019$.)

```{r e016b-chi-np, fig.width=6, fig.height=4}
chi_sq_np_plot(lag_e016b)
```

```{r e016b-chi-lambda, fig.width=6, fig.height=4}
chi_sq_lambda_plot(lag_e016b)
```

```{r e016b-output-T1, fig.width=6, fig.height=4}
rolling_tc_plot(roll_e016b)   # PLE saturates at the tc bound
```

# Reproducibility and the cache

This document separates three tiers of computation:

1. **Live (always run).** The SSE surfaces (§1) and all data simulation — cheap
   ($\approx$ 0.3 s per surface), so they are recomputed on every knit.
2. **Cached single-model fits.** `fit_e008`, `fit_e009` and `spx_fit`
   ($\approx$ 1–3 min each) are cached as `fit_e008.rds`, `fit_e009.rds`,
   `spx_fit.rds`. Knit with `params = list(recompute = TRUE)` to re-run and
   refresh them; the calibration code is shown in full above.
3. **Precomputed multi-fit tables (reload-only).** The 20-level noise experiments
   (`noise_*.rds`, $\approx$ 45 min) and the rolling-window calibrations
   (`lagrange_rolling_*.rds`, tens of minutes to $\approx$ 1 h) are shipped
   precomputed. Their generators are not yet wired into this file (tracked as a
   TODO in `misc/change_list.md`); the specification for each is given in its
   section, so they can be regenerated with `lppls_rolling()` / `fit_lppls()`.

The only input that cannot be regenerated deterministically is the `e009` GARCH
series (`rugarch` is version-sensitive), so it is always loaded from
`e009_garch_series.rds`.

```{r session, echo=FALSE}
sessionInfo()
```
